---
title: 'CAMDA: Classify DILI drugs (Less+Most-Concern-DILI vs. No-Concern-DILI)'
author: "Joaquim Aguirre-Plans"
date: "6/5/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
### Install packages ###
# Install Bioconductor to install "rhdf5", a dependency of cmapR
#install.packages("BiocManager")
#BiocManager::install("rhdf5")
# Install data.table, testthat, dependencies of cmapR
#install.packages("data.table")
#install.packages("testthat")
# cmapR has been installed manually by downloading the package on github (https://github.com/cmap/cmapR) and running the following commands:
# R CMD build cmapR
# R CMD check cmapR_1.0.1.tar.gz
#install.packages("/home/quim/R/x86_64-pc-linux-gnu-library/3.5/cmapR_1.0.1.tar.gz", type="source", repos=NULL)
# Install GGPlot2 and its dependencies
#install.packages('Rcpp', dependencies = TRUE)
#install.packages('ggplot2', dependencies = TRUE)
#install.packages('data.table', dependencies = TRUE)
# Install ggbiplot to plot the PCA
#library(devtools)
#install_github("vqv/ggbiplot")

### Load packages ###
library(cmapR)
library(ggplot2)
library(caret)
```

```{r, include=FALSE}
### Define files ###
#setwd("/sbi/users/interchange/emre/quim/camda")
drugs_file <- "/home/quim/PHD/Projects/camda/camda_data/CAMDA_l1000_1314compounds-pert_iname.rda"
drug_info_file <- "/sbi/users/interchange/emre/quim/camda/CAMDA_l1000_1314compounds-info.rda"
dilirank_file <- "/home/quim/PHD/Projects/camda/camda_data/CAMDA_l1000_1314compounds-dilirank.v2.rda"
expression_file <- "/sbi/users/interchange/emre/quim/camda/CAMDA_l1000_1314compounds-GSE92742_Level5_gct.rda"
landmark_genes_file <- "/home/quim/PHD/Projects/camda/additional_data/GSE92742_Broad_LINCS_gene_info_delta_landmark.txt"
gene_info_file <- "/home/quim/PHD/Projects/camda/additional_data/GSE92742_Broad_LINCS_gene_info.txt"
cell_info_file <- "/home/quim/PHD/Projects/camda/additional_data/GSE92742_Broad_LINCS_cell_info.txt"
```


First we load the R objects needed.

```{r}
load(drugs_file)
load(drug_info_file)
load(dilirank_file)
load(expression_file) # Requires cmapR
gene_info_df <- read.csv(gene_info_file, header=TRUE, sep="\t")
landmark_genes <- gene_info_df$pr_gene_id[gene_info_df$pr_is_lm==1]
```


We subset the DILIConcern data by those drugs that have the same label in the columns DILIConcern and vDILIConcern.

```{r process_dili_info}
# Get independent dataset (Ambiguous drugs)
independent_df <- drank.sel[drank.sel$vDILIConcern == "Ambiguous DILI-concern",]
independent_drugs <- independent_df$pert_iname

# Get cross-validation dataset
drank.sel$DILIConcern[drank.sel$DILIConcern == "No-DILI-concern"] <- "No-DILI-Concern" # Correct the 4 entries with lowercase concern
# We remove the first letter from vDILIConcern entries (except Ambiguous)
drank.sel$vDILIConcern[drank.sel$vDILIConcern != "Ambiguous DILI-concern"] <- sub('.', '', drank.sel$vDILIConcern[drank.sel$vDILIConcern != "Ambiguous DILI-concern"])
# We subset the table by those drugs that have the same value in vDILIConcern and DILIConcern columns
dilirank_df <- drank.sel[drank.sel$vDILIConcern == drank.sel$DILIConcern,]
all_drugs <- dilirank_df$pert_iname

# Remove the outliers
outliers = c('daunorubicin', 'vorinostat')

# Define drugs without outliers
all_drugs_no_outliers <- all_drugs[!all_drugs %in% outliers]

# Check how many drugs there are in the different groups with respect to severity
table(dilirank_df$DILIConcern[dilirank_df$pert_iname%in%all_drugs_no_outliers], dilirank_df$Severity.Class[dilirank_df$pert_iname%in%all_drugs_no_outliers])

# Define different groups of drugs and calculate proportions
most_concern_drugs <- dilirank_df$pert_iname[dilirank_df$pert_iname %in% all_drugs_no_outliers & dilirank_df$DILIConcern == "Most-DILI-Concern"]
less_concern_drugs <- dilirank_df$pert_iname[dilirank_df$pert_iname %in% all_drugs_no_outliers & dilirank_df$DILIConcern == "Less-DILI-Concern"]
no_concern_drugs <- dilirank_df$pert_iname[dilirank_df$pert_iname %in% all_drugs_no_outliers & dilirank_df$DILIConcern == "No-DILI-Concern"]

# Define different groups of drugs based on severity
#drugs_most <- dilirank_df$pert_iname[dilirank_df$pert_iname %in% all_drugs_no_outliers & dilirank_df$Severity.Class>=5] #55
#drugs_mid <- union(dilirank_df$pert_iname[dilirank_df$pert_iname %in% all_drugs_no_outliers & dilirank_df$DILIConcern=="Less-DILI-Concern" & dilirank_df$Severity.Class>=4], dilirank_df$pert_iname[dilirank_df$pert_iname %in% all_drugs_no_outliers & dilirank_df$DILIConcern=="Most-DILI-Concern" & dilirank_df$Severity.Class>=4 & dilirank_df$Severity.Class<8]) #43
#drugs_less <- dilirank_df$pert_iname[dilirank_df$pert_iname %in% all_drugs_no_outliers & dilirank_df$Severity.Class==3] #51
#drugs_no <- dilirank_df$pert_iname[dilirank_df$pert_iname %in% all_drugs_no_outliers & dilirank_df$DILIConcern=="No-DILI-Concern"] #49
#group_most <- union(drugs_most, drugs_no)
#group_mid <- union(drugs_mid, drugs_no)
#group_less <- union(drugs_less, drugs_no)
```


We define the functions to train the machine learning classifiers:

```{r machine_learning}

#### MACHINE LEARNING TRAINING FUNCTION (by cross-validation) ####
train.model<-function(training, testing, type_analysis="discrete", ml.method="rf") {

  # Create train and test set
  #inTrain = createDataPartition(data$cat, p = 0.7, list=F)
  #training <- data[ inTrain,]
  #testing <- data[-inTrain,]
  
  # Define parameters
  ctrl = trainControl(method = "repeatedcv", number = 10, repeats = 3)

  # Train
  print(ml.method)
  modFit = train(cat ~ ., data = training, method = ml.method, trControl = ctrl, verbose=F) 
  pred = predict(modFit, testing)
  #print('PREDICTORS:')
  #print(predictors(modFit))
  #print(modFit)
  if(type_analysis == "discrete") { 
		cor = NA
		a = confusionMatrix(table(pred, testing$cat))
		print('CONFUSION MATRIX:')
    print(a)
  } else {
    # Calculate correlation
    pred <- as.numeric(pred)
    testing$cat <- as.numeric(testing$cat)
    cor = cor(as.numeric(pred), as.numeric(testing$cat))
    # Assign category to testing and predictions
    pred_df<-data.frame(pred)
    pred_df$DILI <- "no"
    if (length(pred_df[pred_df$pred>=2,]$DILI)>0){
      pred_df[pred_df$pred>=2,]$DILI <- "less/most"
    }
    testing$DILI <- "no"
    if (length(testing$cat>=2)){
      testing[testing$cat>=2,]$DILI <- "less/most"
    }
    # Create confusion matrix based on category
    u <- union(pred_df$DILI, testing$DILI)
    t <- table(factor(pred_df$DILI, u), factor(testing$DILI, u))
		a = confusionMatrix(t)
		print('CONFUSION MATRIX:')
		print(a)
		print('TESTING VALUES:')
		print(testing$cat)
		print('PREDICTION VALUES:')
		print(pred)
		print('CORRELATION:')
		print(cor)
  }

  return(list(modFit=modFit, a=a, cor=cor));
}

```


We also define a function to subset the gene expression data from the gct file:

```{r subset_gene_expression}

#### SUBSET EXPRESSION FUNCTION ####
subset.expression<-function(gct, selected_genes, drugs, cell_id="PHH", pert_idose="10 µM", pert_itime="24 h", merge_samples=FALSE, independent=FALSE) {

  # Subset the gct object by cell line, dose, time and genes
  id_selected_samples <- which(gct@cdesc$cell_id == cell_id & gct@cdesc$pert_idose == pert_idose & gct@cdesc$pert_itime == pert_itime)
  id_selected_genes <- which(gct@rid %in% selected_genes)
  gct_subset <- subset.gct(gct, cid=id_selected_samples, rid=id_selected_genes)

  # Subset by all drugs of interest
  id_drugs <- which(gct_subset@cdesc$pert_iname %in% drugs)
  gct_drugs <- subset.gct(gct_subset, cid=id_drugs)
  
  # Create table of expression
  expression_df <- data.frame(t(gct_drugs@mat))
  
  # Get info of DILIrank
  mapping_drugs <- data.frame(gct_drugs@cid, gct_drugs@cdesc$pert_id, gct_drugs@cdesc$pert_iname)
  colnames(mapping_drugs) <- c("cid", "pert_id", "pert_iname")
  if (independent == FALSE){
    mapping_drugs <- merge(x = mapping_drugs, y = dilirank_df[c("pert_iname", "Severity.Class", "DILIConcern")], by = "pert_iname")
  } else {
    mapping_drugs <- merge(x = mapping_drugs, y = independent_df[c("pert_iname", "Severity.Class", "DILIConcern")], by = "pert_iname")
  }
  
  # Get severity values and DILIrank categories and map them to the table of expression
  severity <- c()
  dilirank <- c()
  pert_inames <- c()
  for (cid in rownames(expression_df)){
    row <- mapping_drugs[mapping_drugs$cid==cid,]
    severity[length(severity)+1] <- row$Severity.Class
    dilirank[length(dilirank)+1] <- row$DILIConcern
    pert_inames[length(pert_inames)+1] <- as.character(row$pert_iname)
  }
  expression_df$pert_iname <- pert_inames
  expression_df$severity <- severity
  expression_df$dilirank <- dilirank
  if (merge_samples==TRUE){
    #expression_df <- aggregate(.~pert_iname+severity+dilirank,expression_df,median) # Join samples from same drug
    expression_df <- aggregate(.~pert_iname+severity+dilirank,expression_df,function(val){y <- val[which.max( abs(val) )]; return(y)}) # Join samples from same drug
  }
  return(expression_df);
}


#### SUBSET EXPRESSION FUNCTION (for multiple cell lines) ####
subset.expression.multiple.cells<-function(gct, selected_genes, drugs, cell_ids, pert_idose="10 µM", pert_itime="24 h", merge_samples=FALSE, independent=FALSE) {

  # Subset the gct object by cell line, dose, time and genes
  id_selected_samples <- which(gct@cdesc$cell_id %in% cell_ids & gct@cdesc$pert_idose == pert_idose & gct@cdesc$pert_itime == pert_itime)
  id_selected_genes <- which(gct@rid %in% selected_genes)
  gct_subset <- subset.gct(gct, cid=id_selected_samples, rid=id_selected_genes)

  # Subset by all drugs of interest
  id_drugs <- which(gct_subset@cdesc$pert_iname %in% drugs)
  gct_drugs <- subset.gct(gct_subset, cid=id_drugs)
  
  # Create table of expression
  expression_df <- data.frame(t(gct_drugs@mat))
  
  # Get info of DILIrank
  mapping_drugs <- data.frame(gct_drugs@cid, gct_drugs@cdesc$pert_id, gct_drugs@cdesc$pert_iname, gct_drugs@cdesc$cell_id)
  colnames(mapping_drugs) <- c("cid", "pert_id", "pert_iname", "cell_id")
  if (independent == FALSE){
    mapping_drugs <- merge(x = mapping_drugs, y = dilirank_df[c("pert_iname", "Severity.Class", "DILIConcern")], by = "pert_iname")
  } else {
    mapping_drugs <- merge(x = mapping_drugs, y = independent_df[c("pert_iname", "Severity.Class", "DILIConcern")], by = "pert_iname")
  }
  
  # Get severity values and DILIrank categories and map them to the table of expression
  severity <- c()
  dilirank <- c()
  pert_inames <- c()
  cell_ids <- c()
  for (cid in rownames(expression_df)){
    row <- mapping_drugs[mapping_drugs$cid==cid,]
    severity[length(severity)+1] <- row$Severity.Class
    dilirank[length(dilirank)+1] <- row$DILIConcern
    pert_inames[length(pert_inames)+1] <- as.character(row$pert_iname)
    cell_ids[length(cell_ids)+1] <- as.character(row$cell_id)
  }
  expression_df$pert_iname <- pert_inames
  #expression_df$cell_id <- cell_ids
  expression_df$severity <- severity
  expression_df$dilirank <- dilirank
  if (merge_samples==TRUE){
    #expression_df <- aggregate(.~pert_iname+cell_id+severity+dilirank,expression_df,median) # Join samples from same drug
    expression_df <- aggregate(. ~ pert_iname+severity+dilirank,expression_df,function(val){y <- val[which.max( abs(val) )]; return(y)}) # Join samples from same drug
  }
  return(expression_df);
}

```



## Use landmark genes from Wilcoxon test with p-value > 0.05 (PHH, 10 µM, 24 h)

We will use the genes that where differentiating better Most vs. No DILI drugs when calculating the Wilcoxon test:

```{r wilcox_read}
# Read the wilcoxon file and get the genes
wilcox_file <- "/home/quim/PHD/Projects/camda/results/gene_test_landmark_phh_10_24_mixlessmost.tsv"
wilcox_df = read.csv(wilcox_file, header = TRUE, sep = "\t")
selected_genes <- unique(wilcox_df$gene_id[wilcox_df$p.value<0.05 & wilcox_df$landmark==TRUE])
```

We will subset gene expression data by PHH, 10 µM, 24 h:

```{r wilcox_subset}
# Subset the GCT object by cell ID PHH, 10 µM, 24 h
expression_wilcox_df <- subset.expression(gct, selected_genes, all_drugs_no_outliers, cell_id="PHH", pert_idose="10 µM", pert_itime="24 h", merge_samples = TRUE)
# Subset gene expression for independent drugs as well
expression_wilcox_ind_df <- subset.expression(gct, selected_genes, independent_drugs, cell_id="PHH", pert_idose="10 µM", pert_itime="24 h", merge_samples = TRUE, independent = TRUE)
```

We will prepare the data to train:

```{r wilcox_prepare}

#### PREPARE BALANCED DATASETS FUNCTION ####
prepare.dataset<-function(expression_data, num_datasets, most_concern_drugs, less_concern_drugs, no_concern_drugs, type_analysis="discrete") {
  
  # Define number of drugs
  most_less_concern_drugs <- union(most_concern_drugs, less_concern_drugs)
  prop_most = length(most_concern_drugs) / length(most_less_concern_drugs)
  prop_less = length(less_concern_drugs) / length(most_less_concern_drugs)
  num_most = round(prop_most * length(no_concern_drugs))
  num_less = round(prop_less * length(no_concern_drugs))

  # Get the drugs for the testing dataset
  drugs_testing <- c(sample(less_concern_drugs, round(num_less*0.3), replace=F), sample(most_concern_drugs, round(num_most*0.3), replace=F), sample(no_concern_drugs, round(length(no_concern_drugs)*0.3), replace=F))
  testing <- expression_data[expression_data$pert_iname %in% drugs_testing,]
  testing$pert_iname <- NULL
  if (type_analysis == "discrete"){
    testing$severity <- NULL
    colnames(testing)[match("dilirank", colnames(testing))] <- "cat"
    testing$cat[testing$cat == "No-DILI-Concern"] <- "vNo-DILI-Concern" # Rename No records
    testing$cat[testing$cat == "Less-DILI-Concern" | testing$cat == "Most-DILI-Concern"] <- "vMost-DILI-Concern / vLess-DILI-Concern" # Rename Most/Less records
  } else {
    testing$dilirank <- NULL
    colnames(testing)[match("severity", colnames(testing))] <- "cat"
  }
  
  # Get the drugs for the training datasets
  training_datasets <- list()
  for (i in 1:num_datasets){
    # Get drugs
    drugs_training <- c(sample(less_concern_drugs[!less_concern_drugs%in%drugs_testing], round(num_less*0.7), replace=F), sample(most_concern_drugs[!most_concern_drugs%in%drugs_testing], round(num_most*0.7), replace=F), sample(no_concern_drugs[!no_concern_drugs%in%drugs_testing], round(length(no_concern_drugs)*0.7), replace=F))
    # Get gene expression
    training <- expression_data[expression_data$pert_iname %in% drugs_training,]
    # Prepare table
    training$pert_iname <- NULL
    if (type_analysis == "discrete"){
      training$severity <- NULL
      colnames(training)[match("dilirank", colnames(training))] <- "cat"
      training$cat[training$cat == "No-DILI-Concern"] <- "vNo-DILI-Concern" # Rename No records
      training$cat[training$cat == "Less-DILI-Concern" | training$cat == "Most-DILI-Concern"] <- "vMost-DILI-Concern / vLess-DILI-Concern" # Rename Most/Less records
    } else{
      training$dilirank <- NULL
      colnames(training)[match("severity", colnames(training))] <- "cat"
    }
    training_datasets[[length(training_datasets)+1]] <- training
  }
  return(list(testing=testing, training_datasets=training_datasets));
}
```

We will make a function to train three datasets and combine them:

```{r train_and_combine}
#### TRAIN THREE CLASSIFIERS FOR THREE DATASETS AND COMBINE THEM ####
train.combine.classifiers<-function(datasets.list, ml.method="rf", type_analysis="discrete") {

  # Train 3 models
  train.list.1 <- train.model(training = datasets.list$training_datasets[[1]], testing = datasets.list$testing, type_analysis = type_analysis, ml.method = ml.method)
  train.list.2 <- train.model(training = datasets.list$training_datasets[[2]], testing = datasets.list$testing, type_analysis = type_analysis, ml.method = ml.method)
  train.list.3 <- train.model(training = datasets.list$training_datasets[[3]], testing = datasets.list$testing, type_analysis = type_analysis, ml.method = ml.method)

  # Combine them
  pred.1 = predict(train.list.1$modFit, datasets.list$testing)
  pred.2 = predict(train.list.2$modFit, datasets.list$testing)
  pred.3 = predict(train.list.3$modFit, datasets.list$testing)
  pred.comb = data.frame(pred.1, pred.2, pred.3, cat=datasets.list$testing$cat)
  if (type_analysis=="discrete"){
    modFit = train(cat ~ ., data=pred.comb, method = "rf")
    pred = predict(modFit, datasets.list$testing$cat)
    a = confusionMatrix(table(pred, datasets.list$testing$cat))
    print('COMBINED CONFUSION MATRIX:')
    print(a)
  } else{
    modFit = train(cat ~ ., data=pred.comb, method = "gam")
    pred = predict(modFit, datasets.list$testing$cat)
    pred_df<-data.frame(pred=pred, cat=datasets.list$testing$cat)
    pred_df$DILI <- "no"
    if (length(pred_df[pred_df$pred>=2,]$DILI)>0){
      pred_df[pred_df$pred>=2,]$DILI <- "less/most"
    }
    datasets.list$testing$DILI <- "no"
    if (length(datasets.list$testing$cat[datasets.list$testing$cat>=2])>0){
      datasets.list$testing[datasets.list$testing$cat>=2,]$DILI <- "less/most"
    }
    a = confusionMatrix(table(pred_df$DILI, datasets.list$testing$DILI))
    print('COMBINED CONFUSION MATRIX:')
    print(a)
  }
  return(list(modFit.comb=modFit, modFit.1=train.list.1$modFit, modFit.2=train.list.2$modFit, modFit.3=train.list.3$modFit));
}
```

And a function to test the classifier on an independent dataset:

```{r validate_on_independent}
#### VALIDATE CLASSIFIER ON AN INDEPENDENT DATASET ####
validate.classifier<-function(independent_data, output_file, modFit.comb, modFit.1, modFit.2, modFit.3, type_analysis="discrete") {
  val.data <- independent_data
  val.data$pert_iname <- NULL
  val.data$severity <- NULL
  val.data$dilirank <- NULL

  pred.1 = predict(modFit.1, val.data)
  pred.2 = predict(modFit.2, val.data)
  pred.3 = predict(modFit.3, val.data)
  pred.val.comb = data.frame(pred.1=pred.1, pred.2=pred.2, pred.3=pred.3)
  pred = predict(modFit.comb, pred.val.comb)
  if (type_analysis=="discrete"){
    pred_df <- data.frame(Compound.Name=independent_data$pert_iname, Predicted.Label=pred)
    write.table(pred_df, file = output_file, row.names=FALSE, na="-", col.names=TRUE, sep=",", quote=FALSE)
  } else {
    pred_df <- data.frame(Compound.Name=independent_data$pert_iname, pred=pred)
    pred_df$Predicted.Label <- "vNo-DILI-Concern"
    if (length(pred_df[pred_df$pred>=2,]$Predicted.Label)>0){
      pred_df[pred_df$pred>=2,]$Predicted.Label <- "vMost-DILI-Concern / vLess-DILI-Concern"
    }
    pred_df$pred <- NULL
    write.table(pred_df, file = output_file, row.names=FALSE, na="-", col.names=TRUE, sep=",", quote=FALSE)
  }
  return(pred_df);
}
```

We will train the classifier using discrete category (Most/Less vs. No):

```{r wilcox_train_discrete}
# Get the expression data in an appropriate format
datasets.list <- prepare.dataset(expression_wilcox_df, 3, most_concern_drugs, less_concern_drugs, no_concern_drugs, type_analysis = "discrete")

# Train the classifier by RF
wilcox.discrete.rf.list <- train.combine.classifiers(datasets.list, ml.method="rf", type_analysis="discrete")

# Test the independent dataset (RF)
output_file = "/home/quim/PHD/Projects/camda/camda_data/independent_validation/JAguirre_predictions_mix_wilcox_discrete_rf.txt"
pred_df <- validate.classifier(expression_wilcox_ind_df, output_file, 
                               wilcox.discrete.rf.list$modFit.comb, 
                               wilcox.discrete.rf.list$modFit.1, 
                               wilcox.discrete.rf.list$modFit.2, 
                               wilcox.discrete.rf.list$modFit.3, 
                               type_analysis="discrete")

# Train the classifier by GBM
wilcox.discrete.gbm.list <- train.combine.classifiers(datasets.list, ml.method="gbm", type_analysis="discrete")

# Test the independent dataset (GBM)
output_file = "/home/quim/PHD/Projects/camda/camda_data/independent_validation/JAguirre_predictions_mix_wilcox_discrete_gbm.txt"
pred_df <- validate.classifier(expression_wilcox_ind_df, output_file, 
                               wilcox.discrete.gbm.list$modFit.comb, 
                               wilcox.discrete.gbm.list$modFit.1, 
                               wilcox.discrete.gbm.list$modFit.2, 
                               wilcox.discrete.gbm.list$modFit.3, 
                               type_analysis="discrete")
```

Now, we will try with continuous category (Severity):

```{r wilcox_train_continuous}

# Get the expression data in an appropriate format
datasets.list <- prepare.dataset(expression_wilcox_df, 3, most_concern_drugs, less_concern_drugs, no_concern_drugs, type_analysis = "continuous")

# Train the classifier by RF
wilcox.continuous.rf.list <- train.combine.classifiers(datasets.list, ml.method="rf", type_analysis="continuous")

# Test the independent dataset (RF)
output_file = "/home/quim/PHD/Projects/camda/camda_data/independent_validation/JAguirre_predictions_mix_wilcox_continuous_rf.txt"
pred_df <- validate.classifier(expression_wilcox_ind_df, output_file, 
                               wilcox.continuous.rf.list$modFit.comb, 
                               wilcox.continuous.rf.list$modFit.1, 
                               wilcox.continuous.rf.list$modFit.2, 
                               wilcox.continuous.rf.list$modFit.3, 
                               type_analysis="continuous")

# Train the classifier by GBM
wilcox.continuous.gbm.list <- train.combine.classifiers(datasets.list, ml.method="gbm", type_analysis="continuous")

# Test the independent dataset (GBM)
output_file = "/home/quim/PHD/Projects/camda/camda_data/independent_validation/JAguirre_predictions_mix_wilcox_continuous_gbm.txt"
pred_df <- validate.classifier(expression_wilcox_ind_df, output_file, 
                               wilcox.continuous.gbm.list$modFit.comb, 
                               wilcox.continuous.gbm.list$modFit.1, 
                               wilcox.continuous.gbm.list$modFit.2, 
                               wilcox.continuous.gbm.list$modFit.3, 
                               type_analysis="continuous")
```


## Use landmark genes from Wilcoxon test with p-value > 0.05 (Specific cell lines, 10 µM, 24 h)

We will use the genes that where differentiating better Most vs. No DILI drugs when calculating the Wilcoxon test. We will subset by the following cell lines: 

* A375, FIBRNPC (skin)
* A549, HCC515 (lung)
* BT20, HS578T, MCF10A, MCF7, MDAMB231, SKBR3 (breast)
* HA1E, HEK293T, NKDBA (kidney)
* HEPG2, PHH (liver)
* HT29, SW480 (large intestine)


```{r wilcox_spec_read}
# Read the wilcoxon file and get the genes
wilcox_spec_table <- "/home/quim/PHD/Projects/camda/results/gene_test_landmark_spec_10_24_mixlessmost.tsv"
wilcox_spec_df = read.csv(wilcox_spec_table, header = TRUE, sep = "\t")
selected_genes_spec <- unique(wilcox_spec_df$gene_id[wilcox_spec_df$p.adjust<0.01 & wilcox_spec_df$landmark==TRUE])
```

We will subset gene expression data by PHH, 10 µM, 24 h:

```{r wilcox_spec_subset}
# Subset the GCT object by cell ID PHH, 10 µM, 24 h
cell_ids_selected <- c("A375", "FIBRNPC", "A549", "HCC515", "BT20", "HS578T", "MCF10A", "MCF7", "MDAMB231", "SKBR3", "HA1E", "HEK293T", "NKDBA", "HEPG2", "PHH", "HT29", "SW480")
expression_wilcox_spec_df <- subset.expression.multiple.cells(gct, selected_genes_spec, all_drugs_no_outliers, cell_ids_selected, pert_idose="10 µM", pert_itime="24 h", merge_samples = TRUE)

# Subset gene expression for independent drugs as well
expression_wilcox_spec_ind_df <- subset.expression.multiple.cells(gct, selected_genes_spec, independent_drugs, cell_ids_selected, pert_idose="10 µM", pert_itime="24 h", merge_samples = TRUE, independent = TRUE)

```

We will train the classifier using a discrete category (Most vs. No DILI):

```{r wilcox_spec_train_discrete}

# Get the expression data in an appropriate format
datasets.list <- prepare.dataset(expression_wilcox_spec_df, 3, most_concern_drugs, less_concern_drugs, no_concern_drugs, type_analysis = "discrete")

# Train the classifier by RF
wilcoxspec.discrete.rf.list <- train.combine.classifiers(datasets.list, ml.method="rf", type_analysis="discrete")

# Test the independent dataset (RF)
output_file = "/home/quim/PHD/Projects/camda/camda_data/independent_validation/JAguirre_predictions_mix_wilcoxspec_discrete_rf.txt"
pred_df <- validate.classifier(expression_wilcox_spec_ind_df, output_file, 
                               wilcoxspec.discrete.rf.list$modFit.comb, 
                               wilcoxspec.discrete.rf.list$modFit.1, 
                               wilcoxspec.discrete.rf.list$modFit.2, 
                               wilcoxspec.discrete.rf.list$modFit.3, 
                               type_analysis="discrete")

# Train the classifier by GBM
wilcoxspec.discrete.gbm.list <- train.combine.classifiers(datasets.list, ml.method="gbm", type_analysis="discrete")

# Test the independent dataset (GBM)
output_file = "/home/quim/PHD/Projects/camda/camda_data/independent_validation/JAguirre_predictions_mix_wilcoxspec_discrete_gbm.txt"
pred_df <- validate.classifier(expression_wilcox_spec_ind_df, output_file, 
                               wilcoxspec.discrete.gbm.list$modFit.comb, 
                               wilcoxspec.discrete.gbm.list$modFit.1, 
                               wilcoxspec.discrete.gbm.list$modFit.2, 
                               wilcoxspec.discrete.gbm.list$modFit.3, 
                               type_analysis="discrete")
```

Now, we will try with continuous category (Severity):

```{r wilcox_spec_train_continuous}
# Get the expression data in an appropriate format
datasets.list <- prepare.dataset(expression_wilcox_spec_df, 3, most_concern_drugs, less_concern_drugs, no_concern_drugs, type_analysis = "continuous")

# Train the classifier by RF
wilcoxspec.continuous.rf.list <- train.combine.classifiers(datasets.list, ml.method="rf", type_analysis="continuous")

# Test the independent dataset (RF)
output_file = "/home/quim/PHD/Projects/camda/camda_data/independent_validation/JAguirre_predictions_mix_wilcoxspec_continuous_rf.txt"
pred_df <- validate.classifier(expression_wilcox_spec_ind_df, output_file, 
                               wilcoxspec.continuous.rf.list$modFit.comb, 
                               wilcoxspec.continuous.rf.list$modFit.1, 
                               wilcoxspec.continuous.rf.list$modFit.2, 
                               wilcoxspec.continuous.rf.list$modFit.3, 
                               type_analysis="continuous")

# Train the classifier by GBM
wilcoxspec.continuous.gbm.list <- train.combine.classifiers(datasets.list, ml.method="gbm", type_analysis="continuous")

# Test the independent dataset (GBM)
output_file = "/home/quim/PHD/Projects/camda/camda_data/independent_validation/JAguirre_predictions_mix_wilcoxspec_continuous_gbm.txt"
pred_df <- validate.classifier(expression_wilcox_spec_ind_df, output_file, 
                               wilcoxspec.continuous.gbm.list$modFit.comb, 
                               wilcoxspec.continuous.gbm.list$modFit.1, 
                               wilcoxspec.continuous.gbm.list$modFit.2, 
                               wilcoxspec.continuous.gbm.list$modFit.3, 
                               type_analysis="continuous")
```


## Use Hub Genes (PHH, 10 µM, 24 h)

Let's select the hub genes:

```{r hub_subset}
# Load hub genes
hubgenes_file <- "/home/quim/PHD/Projects/camda/additional_data/PHH_hubgenes.tsv"
hub_df = read.csv(hubgenes_file, header = TRUE, sep = "\t")
hub_genes <- unique(hub_df$entrezID)
#hub_genes_apoptosis <- unique(hub_df$entrezID[!is.na(hub_df$enriched_apoptosis_relTerms)])
#hub_genes_necroptosis <- unique(hub_df$entrezID[!is.na(hub_df$enriched_necroptosis)])
#hub_genes_inflammation <- unique(hub_df$entrezID[!is.na(hub_df$enriched_inflammation)])

# Subset expression
expression_hub_df <- subset.expression(gct, hub_genes, all_drugs_no_outliers, cell_id="PHH", pert_idose="10 µM", pert_itime="24 h", merge_samples = TRUE)

# Subset gene expression for independent drugs as well
expression_hub_ind_df <- subset.expression(gct, hub_genes, independent_drugs, cell_id="PHH", pert_idose="10 µM", pert_itime="24 h", merge_samples = TRUE, independent = TRUE)
```

We will train the classifier using a discrete category (Most vs. No DILI):

```{r all_hub_train_discrete}
# Get the expression data in an appropriate format
datasets.list <- prepare.dataset(expression_hub_df, 3, most_concern_drugs, less_concern_drugs, no_concern_drugs, type_analysis = "discrete")

# Train the classifier by RF
hub.discrete.rf.list <- train.combine.classifiers(datasets.list, ml.method="rf", type_analysis="discrete")

# Train the classifier by GBM
hub.discrete.gbm.list <- train.combine.classifiers(datasets.list, ml.method="gbm", type_analysis="discrete")

# Test the independent dataset
# RF
output_file = "/home/quim/PHD/Projects/camda/camda_data/independent_validation/JAguirre_predictions_mix_hub_discrete_rf.txt"
pred_df <- validate.classifier(expression_hub_ind_df, output_file, 
                               hub.discrete.rf.list$modFit.comb, 
                               hub.discrete.rf.list$modFit.1, 
                               hub.discrete.rf.list$modFit.2, 
                               hub.discrete.rf.list$modFit.3, 
                               type_analysis="discrete")

# GBM
output_file = "/home/quim/PHD/Projects/camda/camda_data/independent_validation/JAguirre_predictions_mix_hub_discrete_gbm.txt"
pred_df <- validate.classifier(expression_hub_ind_df, output_file, 
                               hub.discrete.gbm.list$modFit.comb, 
                               hub.discrete.gbm.list$modFit.1, 
                               hub.discrete.gbm.list$modFit.2, 
                               hub.discrete.gbm.list$modFit.3, 
                               type_analysis="discrete")
```

Now, we will try with continuous category (Severity):

```{r all_hub_train_continuous}

# Get the expression data in an appropriate format
datasets.list <- prepare.dataset(expression_hub_df, 3, most_concern_drugs, less_concern_drugs, no_concern_drugs, type_analysis = "continuous")

# Train the classifier by RF
hub.continuous.rf.list <- train.combine.classifiers(datasets.list, ml.method="rf", type_analysis="continuous")

# Test the independent dataset (RF)
output_file = "/home/quim/PHD/Projects/camda/camda_data/independent_validation/JAguirre_predictions_mix_hub_continuous_rf.txt"
pred_df <- validate.classifier(expression_hub_ind_df, output_file, 
                               hub.continuous.rf.list$modFit.comb, 
                               hub.continuous.rf.list$modFit.1, 
                               hub.continuous.rf.list$modFit.2, 
                               hub.continuous.rf.list$modFit.3, 
                               type_analysis="continuous")

# Train the classifier by GBM
hub.continuous.gbm.list <- train.combine.classifiers(datasets.list, ml.method="gbm", type_analysis="continuous")

# Test the independent dataset (GBM)
output_file = "/home/quim/PHD/Projects/camda/camda_data/independent_validation/JAguirre_predictions_mix_hub_continuous_gbm.txt"
pred_df <- validate.classifier(expression_hub_ind_df, output_file, 
                               hub.continuous.gbm.list$modFit.comb, 
                               hub.continuous.gbm.list$modFit.1, 
                               hub.continuous.gbm.list$modFit.2, 
                               hub.continuous.gbm.list$modFit.3, 
                               type_analysis="continuous")
```


## Use target information (all integrated)

Let's prepare the target data:

```{r target_read}
# Read targets and map them to dilirank info (the dilirank info on targets is wrong!)
targets_file <- "/home/quim/PHD/Projects/camda/additional_data/targets/targets_dgidb_hitpick_sea.tsv"
targets_df <- read.csv(targets_file, header=TRUE, sep="\t", stringsAsFactors=FALSE)
colnames(targets_df)[match("drug", colnames(targets_df))] <- "pert_iname" # Change name to pert_iname
targets_ind_df <- targets_df[targets_df$pert_iname %in% independent_drugs,] # Get independent set
targets_df <- merge(x = targets_df, y = dilirank_df[c("pert_iname", "DILIConcern", "Severity.Class")], by = "pert_iname")
targets_df$DILI <- NULL
targets_df$severity <- NULL
targets_ind_df$DILI <- NULL
targets_ind_df$severity <- NULL
colnames(targets_df)[match("DILIConcern", colnames(targets_df))] <- "dilirank"
colnames(targets_df)[match("Severity.Class", colnames(targets_df))] <- "severity"
targets_df <- targets_df[targets_df$pert_iname %in% all_drugs_no_outliers,]
```

We will train the classifier using a discrete category (Most vs. No DILI):

```{r target_train_discrete}
# Get the expression data in an appropriate format
datasets.list <- prepare.dataset(targets_df, 3, most_concern_drugs, less_concern_drugs, no_concern_drugs, type_analysis = "discrete")

# Train the classifier by RF
targets.discrete.rf.list <- train.combine.classifiers(datasets.list, ml.method="rf", type_analysis="discrete")

# Test the independent dataset (RF)
output_file = "/home/quim/PHD/Projects/camda/camda_data/independent_validation/JAguirre_predictions_mix_targets_discrete_rf.txt"
pred_df <- validate.classifier(targets_ind_df, output_file, 
                               targets.discrete.rf.list$modFit.comb, 
                               targets.discrete.rf.list$modFit.1, 
                               targets.discrete.rf.list$modFit.2, 
                               targets.discrete.rf.list$modFit.3, 
                               type_analysis="discrete")

# Train the classifier by GBM
targets.discrete.gbm.list <- train.combine.classifiers(datasets.list, ml.method="gbm", type_analysis="discrete")

# Test the independent dataset (GBM)
output_file = "/home/quim/PHD/Projects/camda/camda_data/independent_validation/JAguirre_predictions_mix_targets_discrete_gbm.txt"
pred_df <- validate.classifier(targets_ind_df, output_file, 
                               targets.discrete.gbm.list$modFit.comb, 
                               targets.discrete.gbm.list$modFit.1, 
                               targets.discrete.gbm.list$modFit.2, 
                               targets.discrete.gbm.list$modFit.3, 
                               type_analysis="discrete")
```

Now, we will try with continuous category (Severity):

```{r target_train_continuous}

# Get the expression data in an appropriate format
datasets.list <- prepare.dataset(targets_df, 3, most_concern_drugs, less_concern_drugs, no_concern_drugs, type_analysis = "continuous")

# Train the classifier by RF
targets.continuous.rf.list <- train.combine.classifiers(datasets.list, ml.method="rf", type_analysis="continuous")

# Test the independent dataset (RF)
output_file = "/home/quim/PHD/Projects/camda/camda_data/independent_validation/JAguirre_predictions_mix_targets_continuous_rf.txt"
pred_df <- validate.classifier(targets_ind_df, output_file, 
                               targets.continuous.rf.list$modFit.comb, 
                               targets.continuous.rf.list$modFit.1, 
                               targets.continuous.rf.list$modFit.2, 
                               targets.continuous.rf.list$modFit.3, 
                               type_analysis="continuous")

# Train the classifier by GBM
targets.continuous.gbm.list <- train.combine.classifiers(datasets.list, ml.method="gbm", type_analysis="continuous")

# Test the independent dataset (GBM)
output_file = "/home/quim/PHD/Projects/camda/camda_data/independent_validation/JAguirre_predictions_mix_targets_continuous_gbm.txt"
pred_df <- validate.classifier(targets_ind_df, output_file, 
                               targets.continuous.gbm.list$modFit.comb, 
                               targets.continuous.gbm.list$modFit.1, 
                               targets.continuous.gbm.list$modFit.2, 
                               targets.continuous.gbm.list$modFit.3, 
                               type_analysis="continuous")
```



## Use target information (only dgidb)

Let's prepare the target data:

```{r targetdgidb_read}
# # Read targets and map them to dilirank info (the dilirank info on targets is wrong!)
# targets_file <- "/home/quim/PHD/Projects/camda/additional_data/targets/targets_dgidb.tsv"
# targets_df <- read.csv(targets_file, header=TRUE, sep="\t", stringsAsFactors=FALSE)
# colnames(targets_df)[match("drug", colnames(targets_df))] <- "pert_iname" # Change name to pert_iname
# targets_ind_df <- targets_df[targets_df$pert_iname %in% independent_drugs,] # Get independent set
# targets_df <- merge(x = targets_df, y = dilirank_df[c("pert_iname", "DILIConcern", "Severity.Class")], by = "pert_iname")
# targets_df$DILI <- NULL
# targets_df$severity <- NULL
# targets_ind_df$DILI <- NULL
# targets_ind_df$severity <- NULL
# colnames(targets_df)[match("DILIConcern", colnames(targets_df))] <- "dilirank"
# colnames(targets_df)[match("Severity.Class", colnames(targets_df))] <- "severity"
# targets_df <- targets_df[targets_df$pert_iname %in% all_drugs_no_outliers,]
```

We will train the classifier using a discrete category (Most vs. No DILI):

```{r targetdgidb_train_discrete}
# # Get the expression data in an appropriate format
# datasets.list <- prepare.dataset(targets_df, 3, most_concern_drugs, less_concern_drugs, no_concern_drugs, type_analysis = "discrete")
# 
# # Train the classifier by RF
# targetsdgidb.discrete.rf.list <- train.combine.classifiers(datasets.list, ml.method="rf", type_analysis="discrete")
# 
# # Test the independent dataset (RF)
# output_file = "/home/quim/PHD/Projects/camda/camda_data/independent_validation/JAguirre_predictions_mix_targetsdgidb_discrete_rf.txt"
# pred_df <- validate.classifier(targets_ind_df, output_file, 
#                                targetsdgidb.discrete.rf.list$modFit.comb, 
#                                targetsdgidb.discrete.rf.list$modFit.1, 
#                                targetsdgidb.discrete.rf.list$modFit.2, 
#                                targetsdgidb.discrete.rf.list$modFit.3, 
#                                type_analysis="discrete")
# 
# # Train the classifier by GBM
# targetsdgidb.discrete.gbm.list <- train.combine.classifiers(datasets.list, ml.method="gbm", type_analysis="discrete")
# 
# # Test the independent dataset (GBM)
# output_file = "/home/quim/PHD/Projects/camda/camda_data/independent_validation/JAguirre_predictions_mix_targetsdgidb_discrete_gbm.txt"
# pred_df <- validate.classifier(targets_ind_df, output_file, 
#                                targetsdgidb.discrete.gbm.list$modFit.comb, 
#                                targetsdgidb.discrete.gbm.list$modFit.1, 
#                                targetsdgidb.discrete.gbm.list$modFit.2, 
#                                targetsdgidb.discrete.gbm.list$modFit.3, 
#                                type_analysis="discrete")
```

Now, we will try with continuous category (Severity):

```{r targetdgidb_train_continuous}

# # Get the expression data in an appropriate format
# datasets.list <- prepare.dataset(targets_df, 3, most_concern_drugs, less_concern_drugs, no_concern_drugs, type_analysis = "continuous")
# 
# # Train the classifier by RF
# targetsdgidb.continuous.rf.list <- train.combine.classifiers(datasets.list, ml.method="rf", type_analysis="continuous")
# 
# # Test the independent dataset (RF)
# output_file = "/home/quim/PHD/Projects/camda/camda_data/independent_validation/JAguirre_predictions_mix_targets_continuous_rf.txt"
# pred_df <- validate.classifier(targets_ind_df, output_file, 
#                                targetsdgidb.continuous.rf.list$modFit.comb, 
#                                targetsdgidb.continuous.rf.list$modFit.1, 
#                                targetsdgidb.continuous.rf.list$modFit.2, 
#                                targetsdgidb.continuous.rf.list$modFit.3, 
#                                type_analysis="continuous")
# 
# # Train the classifier by GBM
# targetsdgidb.continuous.gbm.list <- train.combine.classifiers(datasets.list, ml.method="gbm", type_analysis="continuous")
# 
# # Test the independent dataset (GBM)
# output_file = "/home/quim/PHD/Projects/camda/camda_data/independent_validation/JAguirre_predictions_mix_targets_continuous_gbm.txt"
# pred_df <- validate.classifier(targets_ind_df, output_file, 
#                                targetsdgidb.continuous.gbm.list$modFit.comb, 
#                                targetsdgidb.continuous.gbm.list$modFit.1, 
#                                targetsdgidb.continuous.gbm.list$modFit.2, 
#                                targetsdgidb.continuous.gbm.list$modFit.3, 
#                                type_analysis="continuous")
```


## Use SMILES information

Let's prepare the SMILES data:

```{r smiles_process}
# Read file and prepare data
tanimoto_file <- "/home/quim/PHD/Projects/camda/additional_data/tanimoto_smiles.tsv"
tanimoto_df <- read.csv(tanimoto_file, header=TRUE, sep="\t", stringsAsFactors=FALSE)
tanimoto_df$DILIConcern[tanimoto_df$DILIConcern == "No-DILI-concern"] <- "No-DILI-Concern" # Correct the 4 entries with lowercase concern
tanimoto_df$pert_iname = rownames(tanimoto_df)
colnames(tanimoto_df)[match("DILIConcern", colnames(tanimoto_df))] <- "dilirank"
tanimoto_df$vDILIConcern <- NULL
tanimoto_ind_df <- tanimoto_df[colnames(tanimoto_df) %in% independent_drugs, rownames(tanimoto_df) %in% all_drugs_no_outliers]
tanimoto_df<-tanimoto_df[colnames(tanimoto_df) %in% all_drugs_no_outliers, rownames(tanimoto_df) %in% all_drugs_no_outliers]
```

We will train the classifier using a discrete category (Most vs. No DILI):

```{r smiles_train_discrete}

# Get the expression data in an appropriate format
datasets.list <- prepare.dataset(tanimoto_df, 3, most_concern_drugs, less_concern_drugs, no_concern_drugs, type_analysis = "discrete")

# Train the classifier by RF
smiles.discrete.rf.list <- train.combine.classifiers(datasets.list, ml.method="rf", type_analysis="discrete")

# Train the classifier by GBM
smiles.discrete.gbm.list <- train.combine.classifiers(datasets.list, ml.method="gbm", type_analysis="discrete")

# Test the independent dataset
# RF
output_file = "/home/quim/PHD/Projects/camda/camda_data/independent_validation/JAguirre_predictions_mix_smiles_discrete_rf.txt"
pred_df <- validate.classifier(tanimoto_ind_df, output_file, 
                               smiles.discrete.rf.list$modFit.comb, 
                               smiles.discrete.rf.list$modFit.1, 
                               smiles.discrete.rf.list$modFit.2, 
                               smiles.discrete.rf.list$modFit.3, 
                               type_analysis="discrete")

# GBM
output_file = "/home/quim/PHD/Projects/camda/camda_data/independent_validation/JAguirre_predictions_mix_smiles_discrete_gbm.txt"
pred_df <- validate.classifier(tanimoto_ind_df, output_file, 
                               smiles.discrete.gbm.list$modFit.comb, 
                               smiles.discrete.gbm.list$modFit.1, 
                               smiles.discrete.gbm.list$modFit.2, 
                               smiles.discrete.gbm.list$modFit.3, 
                               type_analysis="discrete")
```

Now, we will try with continuous category (Severity):

```{r smiles_train_continuous}

# Get the expression data in an appropriate format
datasets.list <- prepare.dataset(tanimoto_df, 3, most_concern_drugs, less_concern_drugs, no_concern_drugs, type_analysis = "continuous")

# Train the classifier by RF
smiles.continuous.rf.list <- train.combine.classifiers(datasets.list, ml.method="rf", type_analysis="continuous")

# Test the independent dataset (RF)
output_file = "/home/quim/PHD/Projects/camda/camda_data/independent_validation/JAguirre_predictions_mix_smiles_continuous_rf.txt"
pred_df <- validate.classifier(tanimoto_ind_df, output_file, 
                               smiles.continuous.rf.list$modFit.comb, 
                               smiles.continuous.rf.list$modFit.1, 
                               smiles.continuous.rf.list$modFit.2, 
                               smiles.continuous.rf.list$modFit.3, 
                               type_analysis="continuous")

# Train the classifier by GBM
smiles.continuous.gbm.list <- train.combine.classifiers(datasets.list, ml.method="gbm", type_analysis="continuous")

# Test the independent dataset (GBM)
output_file = "/home/quim/PHD/Projects/camda/camda_data/independent_validation/JAguirre_predictions_mix_smiles_continuous_gbm.txt"
pred_df <- validate.classifier(tanimoto_ind_df, output_file, 
                               smiles.continuous.gbm.list$modFit.comb, 
                               smiles.continuous.gbm.list$modFit.1, 
                               smiles.continuous.gbm.list$modFit.2, 
                               smiles.continuous.gbm.list$modFit.3, 
                               type_analysis="continuous")
```



## Use DisGeNET Genes (PHH, 10 µM, 24 h)

Let's select the DisGeNET genes:

```{r}
# Read files and get phenotypes
phenotype2gene_file <- "/home/quim/PHD/Projects/camda/guildify_data/phenotype2gene.tsv"
phenotype2gene <- read.csv(phenotype2gene_file, header=TRUE, sep="\t")
phenotypes <- unique(phenotype2gene$diseaseid)
curated_phenotypes <- unique(phenotype2gene$diseaseid[phenotype2gene$source == "CURATED"])
curated_phenotypes <- c("C0023890")
# Get data of phenotypes
for (phenotype in curated_phenotypes){
  curated_gene_ids <-unique(phenotype2gene$geneid[phenotype2gene$diseaseid==phenotype & phenotype2gene$source=="CURATED"])
  name <- unique(phenotype2gene$name[phenotype2gene$diseaseid==phenotype])

  if (length(curated_gene_ids)>10) {
    print(phenotype)
    print(name)
    #print(curated_gene_ids)
  
    # Subset expression
    expression_dg_df <- subset.expression(gct, curated_gene_ids, all_drugs_no_outliers, cell_id="PHH", pert_idose="10 µM", pert_itime="24 h", merge_samples = TRUE)

    # Subset gene expression for independent drugs as well
    expression_dg_ind_df <- subset.expression(gct, curated_gene_ids, independent_drugs, cell_id="PHH", pert_idose="10 µM", pert_itime="24 h", merge_samples = TRUE, independent = TRUE)
    
    #### Train data in using discrete category ####
    
    # Get the expression data in an appropriate format
    datasets.list <- prepare.dataset(expression_dg_df, 3, most_concern_drugs, less_concern_drugs, no_concern_drugs, type_analysis = "discrete")
    
    # Train the classifier by RF
    dg.discrete.rf.list <- train.combine.classifiers(datasets.list, ml.method="rf", type_analysis="discrete")
    
    # Train the classifier by GBM
    dg.discrete.gbm.list <- train.combine.classifiers(datasets.list, ml.method="gbm", type_analysis="discrete")

    # Test the independent dataset (RF)
    output_file = "/home/quim/PHD/Projects/camda/camda_data/independent_validation/JAguirre_predictions_mix_disgenet_discrete_rf.txt"
    pred_df <- validate.classifier(expression_dg_df, output_file, 
                                   dg.discrete.rf.list$modFit.comb, 
                                   dg.discrete.rf.list$modFit.1, 
                                   dg.discrete.rf.list$modFit.2, 
                                   dg.discrete.rf.list$modFit.3, 
                                   type_analysis="discrete")
    
    # Test the independent dataset (GBM)
    output_file = "/home/quim/PHD/Projects/camda/camda_data/independent_validation/JAguirre_predictions_mix_disgenet_discrete_gbm.txt"
    pred_df <- validate.classifier(expression_dg_df, output_file, 
                                   dg.discrete.gbm.list$modFit.comb, 
                                   dg.discrete.gbm.list$modFit.1, 
                                   dg.discrete.gbm.list$modFit.2, 
                                   dg.discrete.gbm.list$modFit.3, 
                                   type_analysis="discrete")
    
    #### Train data using continuous category ####

    # Get the expression data in an appropriate format
    datasets.list <- prepare.dataset(expression_dg_df, 3, most_concern_drugs, less_concern_drugs, no_concern_drugs, type_analysis = "continuous")
    
    # Train the classifier by RF
    dg.continuous.rf.list <- train.combine.classifiers(datasets.list, ml.method="rf", type_analysis="continuous")

    # Test the independent dataset (RF)
    output_file = "/home/quim/PHD/Projects/camda/camda_data/independent_validation/JAguirre_predictions_mix_disgenet_continuous_rf.txt"
    pred_df <- validate.classifier(expression_dg_df, output_file, 
                                   dg.continuous.rf.list$modFit.comb, 
                                   dg.continuous.rf.list$modFit.1, 
                                   dg.continuous.rf.list$modFit.2, 
                                   dg.continuous.rf.list$modFit.3, 
                                   type_analysis="continuous")
    
    # Train the classifier by GBM
    dg.continuous.gbm.list <- train.combine.classifiers(datasets.list, ml.method="gbm", type_analysis="continuous")

    # Test the independent dataset (GBM)
    output_file = "/home/quim/PHD/Projects/camda/camda_data/independent_validation/JAguirre_predictions_mix_disgenet_continuous_gbm.txt"
    pred_df <- validate.classifier(expression_dg_df, output_file, 
                                   dg.continuous.gbm.list$modFit.comb, 
                                   dg.continuous.gbm.list$modFit.1, 
                                   dg.continuous.gbm.list$modFit.2, 
                                   dg.continuous.gbm.list$modFit.3, 
                                   type_analysis="continuous")
  }
}

```



## Use Wilcoxon + Targets information

Let's prepare the target data:

```{r wilcox_target_process}
# Load target data
targets_file <- "/home/quim/PHD/Projects/camda/additional_data/targets/targets_dgidb_hitpick_sea.tsv"
targets_df <- read.csv(targets_file, header=TRUE, sep="\t", stringsAsFactors=FALSE)
colnames(targets_df)[match("drug", colnames(targets_df))] <- "pert_iname" # Change name to pert_iname
targets_ind_df <- targets_df[targets_df$pert_iname %in% independent_drugs,] # Get independent set
targets_df <- merge(x = targets_df, y = dilirank_df[c("pert_iname", "DILIConcern", "Severity.Class")], by = "pert_iname")
targets_df$DILI <- NULL
targets_df$severity <- NULL
targets_ind_df$DILI <- NULL
targets_ind_df$severity <- NULL
colnames(targets_df)[match("DILIConcern", colnames(targets_df))] <- "dilirank"
colnames(targets_df)[match("Severity.Class", colnames(targets_df))] <- "severity"
targets_df <- targets_df[targets_df$pert_iname %in% all_drugs_no_outliers,]


# Load wilcoxon data
wilcox_file <- "/home/quim/PHD/Projects/camda/results/gene_test_landmark_phh_10_24_mixlessmost.tsv"
wilcox_df = read.csv(wilcox_file, header = TRUE, sep = "\t")
selected_genes <- unique(wilcox_df$gene_id[wilcox_df$p.value<0.05 & wilcox_df$landmark==TRUE])

# Subset the GCT object by cell ID PHH, 10 µM, 24 h
expression_wilcox_df <- subset.expression(gct, selected_genes, all_drugs_no_outliers, cell_id="PHH", pert_idose="10 µM", pert_itime="24 h", merge_samples = TRUE)

# Subset gene expression for independent drugs as well
expression_wilcox_ind_df <- subset.expression(gct, selected_genes, independent_drugs, cell_id="PHH", pert_idose="10 µM", pert_itime="24 h", merge_samples = TRUE, independent = TRUE)

# Combine the main datasets
targets_df$dilirank <- NULL
targets_df$severity <- NULL
combined_data <- merge(x = expression_wilcox_df, y = targets_df, by = "pert_iname")

# Combine the independent datasets
combined_data_ind <- merge(x = expression_wilcox_ind_df, y = targets_ind_df, by = "pert_iname")
```


Train the classifiers (discrete):

```{r wilcox_target_train_discrete}

# Get the expression data in an appropriate format
datasets.list <- prepare.dataset(combined_data, 3, most_concern_drugs, less_concern_drugs, no_concern_drugs, type_analysis = "discrete")

# Train the classifier by RF
wilcoxtargets.discrete.rf.list <- train.combine.classifiers(datasets.list, ml.method="rf", type_analysis="discrete")

# Test the independent dataset (RF)
output_file = "/home/quim/PHD/Projects/camda/camda_data/independent_validation/JAguirre_predictions_mix_wilcoxtargets_discrete_rf.txt"
pred_df <- validate.classifier(combined_data_ind, output_file, 
                               wilcoxtargets.discrete.rf.list$modFit.comb, 
                               wilcoxtargets.discrete.rf.list$modFit.1, 
                               wilcoxtargets.discrete.rf.list$modFit.2, 
                               wilcoxtargets.discrete.rf.list$modFit.3, 
                               type_analysis="discrete")

# Train the classifier by GBM
wilcoxtargets.discrete.gbm.list <- train.combine.classifiers(datasets.list, ml.method="gbm", type_analysis="discrete")

# Test the independent dataset (GBM)
output_file = "/home/quim/PHD/Projects/camda/camda_data/independent_validation/JAguirre_predictions_mix_wilcoxtargets_discrete_gbm.txt"
pred_df <- validate.classifier(combined_data_ind, output_file, 
                               wilcoxtargets.discrete.gbm.list$modFit.comb, 
                               wilcoxtargets.discrete.gbm.list$modFit.1, 
                               wilcoxtargets.discrete.gbm.list$modFit.2, 
                               wilcoxtargets.discrete.gbm.list$modFit.3, 
                               type_analysis="discrete")
```

Now, we will try with continuous category (Severity):

```{r wilcox_target_train_continuous}

# Get the expression data in an appropriate format
datasets.list <- prepare.dataset(combined_data, 3, most_concern_drugs, less_concern_drugs, no_concern_drugs, type_analysis = "continuous")

# Train the classifier by RF
wilcoxtargets.continuous.rf.list <- train.combine.classifiers(datasets.list, ml.method="rf", type_analysis="continuous")

# Test the independent dataset (RF)
output_file = "/home/quim/PHD/Projects/camda/camda_data/independent_validation/JAguirre_predictions_mix_wilcoxtargets_continuous_rf.txt"
pred_df <- validate.classifier(combined_data_ind, output_file, 
                               wilcoxtargets.continuous.rf.list$modFit.comb, 
                               wilcoxtargets.continuous.rf.list$modFit.1, 
                               wilcoxtargets.continuous.rf.list$modFit.2, 
                               wilcoxtargets.continuous.rf.list$modFit.3, 
                               type_analysis="continuous")

# Train the classifier by GBM
wilcoxtargets.continuous.gbm.list <- train.combine.classifiers(datasets.list, ml.method="gbm", type_analysis="continuous")

# Test the independent dataset (GBM)
output_file = "/home/quim/PHD/Projects/camda/camda_data/independent_validation/JAguirre_predictions_mix_wilcoxtargets_continuous_gbm.txt"
pred_df <- validate.classifier(combined_data_ind, output_file, 
                               wilcoxtargets.continuous.rf.list$modFit.comb, 
                               wilcoxtargets.continuous.rf.list$modFit.1, 
                               wilcoxtargets.continuous.rf.list$modFit.2, 
                               wilcoxtargets.continuous.rf.list$modFit.3, 
                               type_analysis="continuous")
```



## Use Wilcoxon + SMILES information

Let's prepare the target data:

```{r wilcox_smiles_process}
# Load SMILES data
tanimoto_file <- "/home/quim/PHD/Projects/camda/additional_data/tanimoto_smiles.tsv"
tanimoto_df <- read.csv(tanimoto_file, header=TRUE, sep="\t", stringsAsFactors=FALSE)
tanimoto_df$DILIConcern[tanimoto_df$DILIConcern == "No-DILI-concern"] <- "No-DILI-Concern" # Correct the 4 entries with lowercase concern
tanimoto_df$pert_iname = rownames(tanimoto_df)
colnames(tanimoto_df)[match("DILIConcern", colnames(tanimoto_df))] <- "dilirank"
tanimoto_df$vDILIConcern <- NULL
tanimoto_ind_df <- tanimoto_df[colnames(tanimoto_df) %in% independent_drugs, rownames(tanimoto_df) %in% all_drugs_no_outliers]
tanimoto_df<-tanimoto_df[colnames(tanimoto_df) %in% all_drugs_no_outliers, rownames(tanimoto_df) %in% all_drugs_no_outliers]

# Load wilcoxon data
wilcox_file <- "/home/quim/PHD/Projects/camda/results/gene_test_landmark_phh_10_24_mixlessmost.tsv"
wilcox_df = read.csv(wilcox_file, header = TRUE, sep = "\t")
selected_genes <- unique(wilcox_df$gene_id[wilcox_df$p.value<0.05 & wilcox_df$landmark==TRUE])

# Subset the GCT object by cell ID PHH, 10 µM, 24 h
expression_wilcox_df <- subset.expression(gct, selected_genes, all_drugs_no_outliers, cell_id="PHH", pert_idose="10 µM", pert_itime="24 h", merge_samples = TRUE)

# Subset gene expression for independent drugs as well
expression_wilcox_ind_df <- subset.expression(gct, selected_genes, independent_drugs, cell_id="PHH", pert_idose="10 µM", pert_itime="24 h", merge_samples = TRUE, independent = TRUE)

# Combine the main datasets
tanimoto_df$dilirank <- NULL
tanimoto_df$severity <- NULL
combined_data <- merge(x = expression_wilcox_df, y = tanimoto_df, by = "pert_iname")

# Combine the independent datasets
tanimoto_ind_df$dilirank <- NULL
tanimoto_ind_df$severity <- NULL
combined_data_ind <- merge(x = expression_wilcox_ind_df, y = tanimoto_ind_df, by = "pert_iname")
```


Train the classifier (discrete):

```{r wilcox_smiles_train_discrete}

# Get the expression data in an appropriate format
datasets.list <- prepare.dataset(combined_data, 3, most_concern_drugs, less_concern_drugs, no_concern_drugs, type_analysis = "discrete")

# Train the classifier by RF
wilcoxsmiles.discrete.rf.list <- train.combine.classifiers(datasets.list, ml.method="rf", type_analysis="discrete")

# Train the classifier by GBM
wilcoxsmiles.discrete.gbm.list <- train.combine.classifiers(datasets.list, ml.method="gbm", type_analysis="discrete")

# Test the independent dataset
# RF
output_file = "/home/quim/PHD/Projects/camda/camda_data/independent_validation/JAguirre_predictions_mix_wilcoxsmiles_discrete_rf.txt"
pred_df <- validate.classifier(combined_data_ind, output_file, 
                               wilcoxsmiles.discrete.rf.list$modFit.comb, 
                               wilcoxsmiles.discrete.rf.list$modFit.1, 
                               wilcoxsmiles.discrete.rf.list$modFit.2, 
                               wilcoxsmiles.discrete.rf.list$modFit.3, 
                               type_analysis="discrete")

# GBM
output_file = "/home/quim/PHD/Projects/camda/camda_data/independent_validation/JAguirre_predictions_mix_wilcoxsmiles_discrete_gbm.txt"
pred_df <- validate.classifier(combined_data_ind, output_file, 
                               wilcoxsmiles.discrete.gbm.list$modFit.comb, 
                               wilcoxsmiles.discrete.gbm.list$modFit.1, 
                               wilcoxsmiles.discrete.gbm.list$modFit.2, 
                               wilcoxsmiles.discrete.gbm.list$modFit.3, 
                               type_analysis="discrete")
```


Now, we will try with continuous category (Severity):

```{r wilcox_smiles_train_continuous}

# Get the expression data in an appropriate format
datasets.list <- prepare.dataset(combined_data, 3, most_concern_drugs, less_concern_drugs, no_concern_drugs, type_analysis = "continuous")

# Train the classifier by RF
wilcoxsmiles.continuous.rf.list <- train.combine.classifiers(datasets.list, ml.method="rf", type_analysis="continuous")

# Test the independent dataset (RF)
output_file = "/home/quim/PHD/Projects/camda/camda_data/independent_validation/JAguirre_predictions_mix_wilcoxsmiles_continuous_rf.txt"
pred_df <- validate.classifier(combined_data_ind, output_file, 
                               wilcoxsmiles.continuous.rf.list$modFit.comb, 
                               wilcoxsmiles.continuous.rf.list$modFit.1, 
                               wilcoxsmiles.continuous.rf.list$modFit.2, 
                               wilcoxsmiles.continuous.rf.list$modFit.3, 
                               type_analysis="continuous")

# Train the classifier by GBM
wilcoxsmiles.continuous.gbm.list <- train.combine.classifiers(datasets.list, ml.method="gbm", type_analysis="continuous")

# Test the independent dataset (GBM)
output_file = "/home/quim/PHD/Projects/camda/camda_data/independent_validation/JAguirre_predictions_mix_wilcoxsmiles_continuous_gbm.txt"
pred_df <- validate.classifier(combined_data_ind, output_file, 
                               wilcoxsmiles.continuous.gbm.list$modFit.comb, 
                               wilcoxsmiles.continuous.gbm.list$modFit.1, 
                               wilcoxsmiles.continuous.gbm.list$modFit.2, 
                               wilcoxsmiles.continuous.gbm.list$modFit.3, 
                               type_analysis="continuous")
```



## Use Wilcoxon + Targets + SMILES information

Let's prepare the target data:

```{r wilcox_target_smiles_process}
# Load targets data
targets_file <- "/home/quim/PHD/Projects/camda/additional_data/targets/targets_dgidb_hitpick_sea.tsv"
targets_df <- read.csv(targets_file, header=TRUE, sep="\t", stringsAsFactors=FALSE)
colnames(targets_df)[match("drug", colnames(targets_df))] <- "pert_iname" # Change name to pert_iname
targets_ind_df <- targets_df[targets_df$pert_iname %in% independent_drugs,] # Get independent set
targets_df <- merge(x = targets_df, y = dilirank_df[c("pert_iname", "DILIConcern", "Severity.Class")], by = "pert_iname")
targets_df$DILI <- NULL
targets_df$severity <- NULL
targets_ind_df$DILI <- NULL
targets_ind_df$severity <- NULL
colnames(targets_df)[match("DILIConcern", colnames(targets_df))] <- "dilirank"
colnames(targets_df)[match("Severity.Class", colnames(targets_df))] <- "severity"
targets_df <- targets_df[targets_df$pert_iname %in% all_drugs_no_outliers,]

# Load SMILES data
tanimoto_file <- "/home/quim/PHD/Projects/camda/additional_data/tanimoto_smiles.tsv"
tanimoto_df <- read.csv(tanimoto_file, header=TRUE, sep="\t", stringsAsFactors=FALSE)
tanimoto_df$DILIConcern[tanimoto_df$DILIConcern == "No-DILI-concern"] <- "No-DILI-Concern" # Correct the 4 entries with lowercase concern
tanimoto_df$pert_iname = rownames(tanimoto_df)
colnames(tanimoto_df)[match("DILIConcern", colnames(tanimoto_df))] <- "dilirank"
tanimoto_df$vDILIConcern <- NULL
tanimoto_ind_df <- tanimoto_df[colnames(tanimoto_df) %in% independent_drugs, rownames(tanimoto_df) %in% all_drugs_no_outliers]
tanimoto_df<-tanimoto_df[colnames(tanimoto_df) %in% all_drugs_no_outliers, rownames(tanimoto_df) %in% all_drugs_no_outliers]

# Load wilcoxon data
wilcox_file <- "/home/quim/PHD/Projects/camda/results/gene_test_landmark_phh_10_24_mixlessmost.tsv"
wilcox_df = read.csv(wilcox_file, header = TRUE, sep = "\t")
selected_genes <- unique(wilcox_df$gene_id[wilcox_df$p.value<0.05 & wilcox_df$landmark==TRUE])
# Subset the GCT object by cell ID PHH, 10 µM, 24 h
expression_wilcox_df <- subset.expression(gct, selected_genes, all_drugs_no_outliers, cell_id="PHH", pert_idose="10 µM", pert_itime="24 h", merge_samples = TRUE)
# Subset gene expression for independent drugs as well
expression_wilcox_ind_df <- subset.expression(gct, selected_genes, independent_drugs, cell_id="PHH", pert_idose="10 µM", pert_itime="24 h", merge_samples = TRUE, independent = TRUE)

# Combine the main datasets
targets_df$dilirank <- NULL
targets_df$severity <- NULL
tanimoto_df$dilirank <- NULL
tanimoto_df$severity <- NULL
combined_data1 <- merge(x = expression_wilcox_df, y = targets_df, by = "pert_iname")
combined_data2 <- merge(x = combined_data1, y = tanimoto_df, by = "pert_iname")

# Combine the independent datasets
tanimoto_ind_df$dilirank <- NULL
tanimoto_ind_df$severity <- NULL
combined_data_ind1 <- merge(x = expression_wilcox_ind_df, y = targets_ind_df, by = "pert_iname")
combined_data_ind2 <- merge(x = combined_data_ind1, y = tanimoto_ind_df, by = "pert_iname")
```


Train the classifier (discrete):

```{r wilcox_target_smiles_train_discrete}
# Get the expression data in an appropriate format
datasets.list <- prepare.dataset(combined_data2, 3, most_concern_drugs, less_concern_drugs, no_concern_drugs, type_analysis = "discrete")

# Train the classifier by RF
wilcoxtargetsmiles.discrete.rf.list <- train.combine.classifiers(datasets.list, ml.method="rf", type_analysis="discrete")

# Test the independent dataset (RF)
output_file = "/home/quim/PHD/Projects/camda/camda_data/independent_validation/JAguirre_predictions_mix_wilcoxtargetsmiles_discrete_rf.txt"
pred_df <- validate.classifier(combined_data_ind2, output_file, 
                               wilcoxtargetsmiles.discrete.rf.list$modFit.comb, 
                               wilcoxtargetsmiles.discrete.rf.list$modFit.1, 
                               wilcoxtargetsmiles.discrete.rf.list$modFit.2, 
                               wilcoxtargetsmiles.discrete.rf.list$modFit.3, 
                               type_analysis="discrete")

# Train the classifier by GBM
wilcoxtargetsmiles.discrete.gbm.list <- train.combine.classifiers(datasets.list, ml.method="gbm", type_analysis="discrete")

# Test the independent dataset (GBM)
output_file = "/home/quim/PHD/Projects/camda/camda_data/independent_validation/JAguirre_predictions_mix_wilcoxtargetsmiles_discrete_gbm.txt"
pred_df <- validate.classifier(combined_data_ind2, output_file, 
                               wilcoxtargetsmiles.discrete.gbm.list$modFit.comb, 
                               wilcoxtargetsmiles.discrete.gbm.list$modFit.1, 
                               wilcoxtargetsmiles.discrete.gbm.list$modFit.2, 
                               wilcoxtargetsmiles.discrete.gbm.list$modFit.3, 
                               type_analysis="discrete")
```


Now, we will try with continuous category (Severity):

```{r wilcox_target_smiles_train_continuous}

# Get the expression data in an appropriate format
datasets.list <- prepare.dataset(combined_data2, 3, most_concern_drugs, less_concern_drugs, no_concern_drugs, type_analysis = "continuous")

# Train the classifier by RF
wilcoxtargetsmiles.continuous.rf.list <- train.combine.classifiers(datasets.list, ml.method="rf", type_analysis="continuous")

# Test the independent dataset (RF)
output_file = "/home/quim/PHD/Projects/camda/camda_data/independent_validation/JAguirre_predictions_mix_wilcoxtargetsmiles_continuous_rf.txt"
pred_df <- validate.classifier(combined_data_ind2, output_file, 
                               wilcoxtargetsmiles.continuous.rf.list$modFit.comb, 
                               wilcoxtargetsmiles.continuous.rf.list$modFit.1, 
                               wilcoxtargetsmiles.continuous.rf.list$modFit.2, 
                               wilcoxtargetsmiles.continuous.rf.list$modFit.3, 
                               type_analysis="continuous")

# Train the classifier by GBM
wilcoxtargetsmiles.continuous.gbm.list <- train.combine.classifiers(datasets.list, ml.method="gbm", type_analysis="continuous")

# Test the independent dataset (GBM)
output_file = "/home/quim/PHD/Projects/camda/camda_data/independent_validation/JAguirre_predictions_mix_wilcoxtargetsmiles_continuous_gbm.txt"
pred_df <- validate.classifier(combined_data_ind2, output_file, 
                               wilcoxtargetsmiles.continuous.gbm.list$modFit.comb, 
                               wilcoxtargetsmiles.continuous.gbm.list$modFit.1, 
                               wilcoxtargetsmiles.continuous.gbm.list$modFit.2, 
                               wilcoxtargetsmiles.continuous.gbm.list$modFit.3, 
                               type_analysis="continuous")
```

