---
title: "CMap Drug Safety Challenge: Analysis of compounds"
author: "Joaquim Aguirre-Plans"
date: "March 28, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
### Install packages ###
# Install Bioconductor to install "rhdf5", a dependency of cmapR
#install.packages("BiocManager")
#BiocManager::install("rhdf5", version = "3.8")
# cmapR has been installed manually by downloading the package on github (https://github.com/cmap/cmapR) and running the following commands:
# R CMD build cmapR
# R CMD check cmapR_1.0.1.tar.gz
#install.packages("/home/quim/R/x86_64-pc-linux-gnu-library/3.5/cmapR_1.0.1.tar.gz", type="source", repos=NULL)
# Install GGPlot2 and its dependencies
#install.packages('Rcpp', dependencies = TRUE)
#install.packages('ggplot2', dependencies = TRUE)
#install.packages('data.table', dependencies = TRUE)
# Install ggbiplot to plot the PCA
#library(devtools)
#install_github("vqv/ggbiplot")

### Load packages ###
library(cmapR)
library(ggplot2)
#library(ggbiplot)
```

```{r, include=FALSE}
### Define files ###
#setwd("/sbi/users/interchange/emre/quim/camda")
place = "home" #home or work
if (place=="work"){
  main_directory = "/home/quim/PHD/Projects/camda"
  bigmem_directory = "/sbi/users/interchange/emre/quim/camda"
} else {
  main_directory = "/Users/quim/Dropbox/UPF/PhD/Projects/camda"
  bigmem_directory = "/Users/quim/Documents/DATA/camda"
}
drugs_file <- paste(main_directory, "camda_data/CAMDA_l1000_1314compounds-pert_iname.rda", sep="/")
drug_info_file <- paste(main_directory, "camda_data/CAMDA_l1000_1314compounds-info.rda", sep="/")
dilirank_file <- paste(main_directory, "camda_data/CAMDA_l1000_1314compounds-dilirank.v2.rda", sep="/")
expression_file <- paste(bigmem_directory, "CAMDA_l1000_1314compounds-GSE92742_Level5_gct.rda", sep="/")
gene_info_file <- paste(main_directory, "additional_data/GSE92742_Broad_LINCS_gene_info.txt", sep="/")
```


## Load files

First I load the R objects needed.

```{r}
load(drugs_file)
load(drug_info_file)
load(dilirank_file)
load(expression_file) # Requires cmapR
```

We load the gene information file as well, containing the landmark genes [More info](https://clue.io/connectopedia/l1000_gene_space).

```{r}
gene_info_df <- read.csv(gene_info_file, header=TRUE, sep="\t")
landmark_genes <- gene_info_df$pr_gene_id[gene_info_df$pr_is_lm==1]
```

There are compounds that are not consistently labelled in the DILIConcern and vDILIConcern columns:

```{r}
drank.sel$DILIConcern[drank.sel$DILIConcern == "No-DILI-concern"] <- "No-DILI-Concern" # Correct the 4 entries with lowercase concern
table(drank.sel$vDILIConcern, drank.sel$DILIConcern)
```

Therefore, we will only consider those drugs labelled consistently:

```{r}
# We remove the first letter from vDILIConcern entries (except Ambiguous)
drank.sel$vDILIConcern[drank.sel$vDILIConcern != "Ambiguous DILI-concern"] <- sub('.', '', drank.sel$vDILIConcern[drank.sel$vDILIConcern != "Ambiguous DILI-concern"])
# We subset the table by those drugs that have the same value in vDILIConcern and DILIConcern columns
dilirank_df <- drank.sel[drank.sel$vDILIConcern == drank.sel$DILIConcern,]
table(dilirank_df$vDILIConcern, dilirank_df$DILIConcern)
```

We have `r length(dilirank_df$pert_iname)` compounds:

* Most-DILI-Concern: `r length(dilirank_df$pert_iname[dilirank_df$DILIConcern == "Most-DILI-Concern"])`
* Less-DILI-Concern: `r length(dilirank_df$pert_iname[dilirank_df$DILIConcern == "Less-DILI-Concern"])`
* No-DILI-Concern: `r length(dilirank_df$pert_iname[dilirank_df$DILIConcern == "No-DILI-Concern"])`


## Analyse the conditions of the compounds

There are three relevant parameters if we want to subset the gene expression data of a given drug:

* cell_id: The type of cell line.
* pert_idose: The quantity of drug dose.
* pert_itime: The duration of drug dose.

We print the different conditions of our DILI-related compounds in a table:

```{r}
# Create new dataframe to store gene expression values for pca
cols <- c("pert_iname", "cell_id", "DILIConcern", "num_doses", "min_dose", "max_dose", "num_times", "min_time", "max_time")
compounds_df <- data.frame(matrix(ncol = length(cols), nrow=0))
colnames(compounds_df) <- cols

# Iterate over each drug and check the conditions
drugs <- dilirank_df$pert_iname
for (drug in drugs){
  dili <- dilirank_df$DILIConcern[dilirank_df$pert_iname == drug]
  idx <- which(gct@cdesc$pert_iname==drug)
  x <- subset.gct(gct, cid=idx) # Subset by drug
  cell_ids <- unique(x@cdesc$cell_id)
  for (cell_id in cell_ids){
    idx_c <- which(x@cdesc$cell_id==cell_id)
    x_c <- subset.gct(x, cid=idx_c) # Subset by cell ID
    doses <- as.numeric(unique(as.numeric(unique(x_c@cdesc$pert_dose[x_c@cdesc$pert_dose_unit == 'µM']))))
    times <- as.numeric(unique(x_c@cdesc$pert_time[x_c@cdesc$pert_time_unit == 'h']))
    num_doses <- length(doses)
    min_dose <- min(doses)
    max_dose <- max(doses)
    num_times <- length(times)
    min_time <- min(times)
    max_time <- max(times)
    doses_string <- paste(doses, collapse=',')
    times_string <- paste(times, collapse=',')
    if (length(doses)>=3 | any(doses>=10)){
      compounds_df[nrow(compounds_df)+1,] <- c(drug, cell_id, dili, num_doses, min_dose, max_dose, num_times, min_time, max_time)
    }
  }
}

output_table <- "/home/quim/PHD/Projects/camda/camda_data/summary_compounds.tsv"
write.table(compounds_df, file = output_table,row.names=FALSE, na="-",col.names=TRUE, sep="\t")
head(compounds_df)
```


## Get compounds with more than three doses

We will select the compounds that have more than three doses. We will print them in a new dataframe. We will also create another dataframe to store the expression values of the selected compounds.

```{r}
# Create new dataframe to store the results
cols <- c("pert_iname", "DILIConcern", "num_doses", "min_dose", "max_dose", "num_times", "min_time", "max_time", "num_samples")
morethan3_df <- data.frame(matrix(ncol = length(cols), nrow=0))
colnames(morethan3_df) <- cols

# Create new dataframe to store the gene expression values
#morethan3matrix <- data.frame(matrix(ncol = 0, nrow=length(gct@rid)))
#rownames(morethan3matrix) <- gct@rid
morethan3matrix <- data.frame(1:length(gct@rid), gct@rid)
colnames(morethan3matrix)<- c("id", "pr_gene_id")
morethan3matrix <- merge(morethan3matrix, gene_info_df[,c("pr_gene_id", "pr_gene_symbol", "pr_is_lm")], by="pr_gene_id")

# When merging, we lose the original order (needed to add faster the expression values). But we keep the order in the "id" column, so that we can recover it again
morethan3matrix <- morethan3matrix[order(morethan3matrix$id), ]
morethan3matrix$id <- NULL

morethan3matrixHEPG2 <- morethan3matrix
morethan3matrixPHH <- morethan3matrix

# Iterate over each drug and check the conditions
drugs <- sort(dilirank_df$pert_iname)
cols <- colnames(morethan3matrix)
colsHEPG2 <- colnames(morethan3matrix)
colsPHH <- colnames(morethan3matrix)
for (drug in drugs){
  dili <- dilirank_df$DILIConcern[dilirank_df$pert_iname == drug]
  idx <- which(gct@cdesc$pert_iname==drug)
  x <- subset.gct(gct, cid=idx) # Subset by drug
  doses <- as.numeric(unique(x@cdesc$pert_dose[x@cdesc$pert_dose_unit == 'µM']))
  times <- as.numeric(unique(x@cdesc$pert_time[x@cdesc$pert_time_unit == 'h']))
  num_doses <- length(doses)
  min_dose <- min(doses)
  max_dose <- max(doses)
  num_times <- length(times)
  min_time <- min(times)
  max_time <- max(times)
  num_samples <- length(x@cid)
  if (length(doses)>=3 & any(doses>=10) & any(times>=24)){
    morethan3_df[nrow(morethan3_df)+1,] <- c(drug, dili, num_doses, min_dose, max_dose, num_times, min_time, max_time, num_samples)
    # Assign a new name to all the samples and keep the expression in the table morethan3matrix
    prev_sample_id = ''
    num <- 1
    for (cid in x@cid){
      cell_id <- x@cdesc$cell_id[x@cid==cid]
      dose <- gsub(" ", "", x@cdesc$pert_idose[x@cid==cid], fixed = TRUE)
      time <- gsub(" ", "", x@cdesc$pert_itime[x@cid==cid], fixed = TRUE)
      #sample_id <- paste(drug, cell_id, dose, time, num, sep='_')
      sample_id <- paste(drug, cell_id, dose, time, sep='_')
      #if (prev_sample_id==sample_id){
      #  num <- num+1
      #  sample_id <- paste(drug, cell_id, dose, time, num, sep='_')
      #}else{
      #  num <- 1
      #  sample_id <- paste(drug, cell_id, dose, time, num, sep='_')
      #}
      prev_sample_id <- sample_id
      #print(sample_id)
      expression <- x@mat[x@cid==cid]
      morethan3matrix[,length(cols)+1] <- unlist(expression)
      if (cell_id=="HEPG2"){
        morethan3matrixHEPG2[,length(colsHEPG2)+1] <- unlist(expression)
        colsHEPG2[length(colsHEPG2)+1] <- sample_id
        print(sample_id)
      } else if (cell_id=="PHH"){
        morethan3matrixPHH[,length(colsPHH)+1] <- unlist(expression)
        colsPHH[length(colsPHH)+1] <- sample_id
        print(sample_id)
      }
      cols[length(cols)+1] <- sample_id
    }
  }
}

# This file will inform about the type of samples in each compound
output_table <- "/home/quim/PHD/Projects/camda/camda_data/summary_compounds_morethan3.tsv"
write.table(morethan3_df, file = output_table,row.names=FALSE, na="-",col.names=TRUE, sep="\t")
head(morethan3_df)

# This file contains the expression values for the samples of the compounds selected (with more than 3 doses) (requested by Giulia)
output_table <- "/home/quim/PHD/Projects/camda/camda_data/matrix_compounds_morethan3_new.tsv"
colnames(morethan3matrix)<- cols
write.table(morethan3matrix, file = output_table,row.names=FALSE, na="-",col.names=TRUE, sep="\t")
morethan3matrix <- read.csv(output_table, header=TRUE, sep="\t")# Read the matrix

# This file contains the expression values for the samples of the compounds selected (with more than 3 doses) for HEPG2 (requested by Giulia)
output_table <- "/home/quim/PHD/Projects/camda/camda_data/matrix_compounds_morethan3_HEPG2_new.tsv"
colnames(morethan3matrixHEPG2)<- colsHEPG2
write.table(morethan3matrixHEPG2, file = output_table,row.names=FALSE, na="-",col.names=TRUE, sep="\t")
morethan3matrixHEPG2 <- read.csv(output_table, header=TRUE, sep="\t")# Read the matrix

# This file contains the expression values for the samples of the compounds selected (with more than 3 doses) for PHH (requested by Giulia)
output_table <- "/home/quim/PHD/Projects/camda/camda_data/matrix_compounds_morethan3_PHH_new.tsv"
colnames(morethan3matrixPHH)<- colsPHH
write.table(morethan3matrixPHH, file = output_table,row.names=FALSE, na="-",col.names=TRUE, sep="\t")
morethan3matrixPHH <- read.csv(output_table, header=TRUE, sep="\t")# Read the matrix
```

