---
title: "CMap Drug Safety Challenge: Gene test across samples of PHH, 10 µM and 24 h"
author: "Joaquim Aguirre-Plans"
date: "April 29, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
### Install packages ###
# Install Bioconductor to install "rhdf5", a dependency of cmapR
#install.packages("BiocManager")
#BiocManager::install("rhdf5", version = "3.8")
# cmapR has been installed manually by downloading the package on github (https://github.com/cmap/cmapR) and running the following commands:
# R CMD build cmapR
# R CMD check cmapR_1.0.1.tar.gz
#install.packages("/home/quim/R/x86_64-pc-linux-gnu-library/3.5/cmapR_1.0.1.tar.gz", type="source", repos=NULL)
# Install GGPlot2 and its dependencies
#install.packages('Rcpp', dependencies = TRUE)
#install.packages('ggplot2', dependencies = TRUE)
#install.packages('data.table', dependencies = TRUE)
# Install ggbiplot to plot the PCA
#library(devtools)
#install_github("vqv/ggbiplot")

### Load packages ###
library(cmapR)
library(ggplot2)
library(ComplexHeatmap) # For Heatmap()
library(circlize) # For Heatmap()
```

```{r, include=FALSE}
### Define files ###
#setwd("/sbi/users/interchange/emre/quim/camda")
drugs_file <- "/home/quim/PHD/Projects/camda/camda_data/CAMDA_l1000_1314compounds-pert_iname.rda"
drug_info_file <- "/sbi/users/interchange/emre/quim/camda/CAMDA_l1000_1314compounds-info.rda"
dilirank_file <- "/home/quim/PHD/Projects/camda/camda_data/CAMDA_l1000_1314compounds-dilirank.v2.rda"
expression_file <- "/sbi/users/interchange/emre/quim/camda/CAMDA_l1000_1314compounds-GSE92742_Level5_gct.rda"
landmark_genes_file <- "/home/quim/PHD/Projects/camda/additional_data/GSE92742_Broad_LINCS_gene_info_delta_landmark.txt"
gene_info_file <- "/home/quim/PHD/Projects/camda/additional_data/GSE92742_Broad_LINCS_gene_info.txt"
cell_info_file <- "/home/quim/PHD/Projects/camda/additional_data/GSE92742_Broad_LINCS_cell_info.txt"
phenotype2gene_file <- "/home/quim/PHD/Projects/camda/guildify_data/phenotype2gene.tsv"
```


## Load files

First we load the R objects needed.

```{r}
load(drugs_file)
load(drug_info_file)
load(dilirank_file)
load(expression_file) # Requires cmapR
```

We load the gene information file as well, containing the landmark genes [More info](https://clue.io/connectopedia/l1000_gene_space).

```{r}
gene_info_df <- read.csv(gene_info_file, header=TRUE, sep="\t")
landmark_genes <- gene_info_df$pr_gene_id[gene_info_df$pr_is_lm==1]
```

I subset the DILIConcern data by those drugs that have the same label in the columns DILIConcern and vDILIConcern.

```{r}
drank.sel$DILIConcern[drank.sel$DILIConcern == "No-DILI-concern"] <- "No-DILI-Concern" # Correct the 4 entries with lowercase concern
# We remove the first letter from vDILIConcern entries (except Ambiguous)
drank.sel$vDILIConcern[drank.sel$vDILIConcern != "Ambiguous DILI-concern"] <- sub('.', '', drank.sel$vDILIConcern[drank.sel$vDILIConcern != "Ambiguous DILI-concern"])
# We subset the table by those drugs that have the same value in vDILIConcern and DILIConcern columns
dilirank_df <- drank.sel[drank.sel$vDILIConcern == drank.sel$DILIConcern,]
most_concern_drugs <- dilirank_df$pert_iname[dilirank_df$DILIConcern == "Most-DILI-Concern"]
less_concern_drugs <- dilirank_df$pert_iname[dilirank_df$DILIConcern == "Less-DILI-Concern"]
most_less_concern_drugs <- union(most_concern_drugs, less_concern_drugs)
no_concern_drugs <- dilirank_df$pert_iname[dilirank_df$DILIConcern == "No-DILI-Concern"]
```

We have `r length(dilirank_df$pert_iname)` compounds:

* Most-DILI-Concern: `r length(most_concern_drugs)`
* Less-DILI-Concern: `r length(less_concern_drugs)`
* No-DILI-Concern: `r length(no_concern_drugs)`

```{r}
# Subset the GCT object by Most-DILI-Concern drugs
id_most <- which(gct@cdesc$pert_iname %in% most_less_concern_drugs)
gct_most <- subset.gct(gct, cid=id_most)
# Subset the GCT object by No-DILI-Concern drugs
id_no <- which(gct@cdesc$pert_iname %in% no_concern_drugs)
gct_no <- subset.gct(gct, cid=id_no)
```

The best conditions to analyse the expression of drugs are 10 µM and 24 h.
We will check the number of samples per type of drug that we have in each cell line:

```{r}
id_most_10_24 <- which(gct_most@cdesc$pert_idose == "10 µM" & gct_most@cdesc$pert_itime == "24 h")
gct_most_10_24 <- subset.gct(gct_most, cid=id_most_10_24)
table(gct_most_10_24@cdesc$cell_id)
id_no_10_24 <- which(gct_no@cdesc$pert_idose == "10 µM" & gct_no@cdesc$pert_itime == "24 h")
gct_no_10_24 <- subset.gct(gct_no, cid=id_no_10_24)
table(gct_no_10_24@cdesc$cell_id)
```

Among these cell lines, there are three liver cell lines according to the cell info file:

```{r}
cell_info_df <- read.csv(cell_info_file, header=TRUE, sep="\t")
cell_info_df[cell_info_df$primary_site == "liver",]
```

The PHH cell line is appropriate, as the cells are primary liver cells. We have 44 samples in Most-DILI-Concern drugs and 64 in No-DILI-Concern drugs. 

```{r}
id_most_phh_10_24 <- which(gct_most_10_24@cdesc$cell_id == "PHH")
gct_most_phh_10_24 <- subset.gct(gct_most_10_24, cid=id_most_phh_10_24)
id_no_phh_10_24 <- which(gct_no_10_24@cdesc$cell_id == "PHH")
gct_no_phh_10_24 <- subset.gct(gct_no_10_24, cid=id_no_phh_10_24)
```


## Compare the expression of each gene across samples of Most-DILI-Concern vs. No-DILI-Concern in PHH cell line, 10 µM and 24 h

For each gene, we will check the difference of expression between Most-DILI-Concern drugs and No-DILI-Concern drugs using a Wilcoxon rank-sum test (or Mann-Whitney U test) ([More info](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/wilcox.test.html)).

```{r}
# Create new dataframe to store the results
cols <- c("gene_id", "landmark", "statistic", "p.value")
wilcox_df <- data.frame(matrix(ncol = length(cols), nrow=0))
colnames(wilcox_df) <- cols

# Calculate the wilcoxon for each gene and store the results
#output_table <- "/home/quim/PHD/Projects/camda/results/gene_test_phh_10_24.tsv"
output_table <- "/home/quim/PHD/Projects/camda/results/gene_test_landmark_phh_10_24_mixlessmost.tsv"
if(!file.exists(output_table)){
  #for (gene_id in gct_most_phh_10_24@rid) {
  for (gene_id in landmark_genes) {
    most_expr <- gct_most_phh_10_24@mat[gct_most_phh_10_24@rid == gene_id]
    no_expr <- gct_no_phh_10_24@mat[gct_no_phh_10_24@rid == gene_id]
    res <- wilcox.test(most_expr, no_expr, alternative = "two.sided", paired = FALSE)
    landmark <- gene_id %in% landmark_genes
    wilcox_df[nrow(wilcox_df)+1,] <- c(gene_id, landmark, res$statistic, res$p.value)
  }
  wilcox_df$p.value <- as.numeric(wilcox_df$p.value)
  wilcox_df$p.adjust <- p.adjust(wilcox_df$p.value, "BH")
  write.table(wilcox_df, file = output_table,row.names=FALSE, na="-",col.names=TRUE, sep="\t")
}else{
  wilcox_df = read.csv(output_table, header = TRUE, sep = "\t")
}
head(wilcox_df)
```

Among the `r length(wilcox_df$gene_id)` genes, there are `r length(wilcox_df$gene_id[wilcox_df$p.adjust<0.05])` genes with an adjusted p-value under 0.05 and `r length(wilcox_df$gene_id[wilcox_df$p.adjust<0.01])` under 0.01. 

Focusing only on the `r length(wilcox_df$gene_id[wilcox_df$landmark==TRUE])` landmark genes, there are `r length(wilcox_df$gene_id[wilcox_df$p.adjust<0.05 & wilcox_df$landmark==TRUE])` genes with a p-value under 0.05 and `r length(wilcox_df$gene_id[wilcox_df$p.adjust<0.01 & wilcox_df$landmark==TRUE])` under 0.01.

We will make a heatmap to plot the expression of the genes for each drug.

```{r}
# Define the genes found
all_wilcox_gene_ids <- unique(wilcox_df$gene_id[wilcox_df$p.value<0.01])
all_wilcox_gene_ids_05 <- unique(wilcox_df$gene_id[wilcox_df$p.value<0.05])
landmark_wilcox_gene_ids <- unique(wilcox_df$gene_id[wilcox_df$p.value<0.01 & wilcox_df$landmark==TRUE])
landmark_wilcox_gene_ids_05 <- unique(wilcox_df$gene_id[wilcox_df$p.value<0.05 & wilcox_df$landmark==TRUE])

# Get a vector of gene symbols ordered equal to the vector of gene IDs
genes_selected <- landmark_wilcox_gene_ids_05
genes_df <- data.frame(1:length(genes_selected), genes_selected)
colnames(genes_df)<- c("id", "pr_gene_id")
genes_df <- merge(genes_df, gene_info_df[,c("pr_gene_id", "pr_gene_symbol")], by="pr_gene_id")
genes_df <- genes_df[order(genes_df$id), ]
genes_df$id <- NULL

# Create new dataframe to store the matrix
cols <- c(most_concern_drugs, no_concern_drugs)
rows <- genes_df$pr_gene_symbol
heat_df <- data.frame(matrix(ncol = length(cols), nrow=length(rows)))
colnames(heat_df) <- cols
rownames(heat_df) <- rows
gene_ids <- genes_df$pr_gene_id

# Introduce the data in the matrix
for (drug in most_concern_drugs){
  id_drug <- which(gct_most_phh_10_24@cdesc$pert_iname == drug)
  id_genes <- which(gct_most_phh_10_24@rid %in% gene_ids)
  gct_drug <- subset.gct(gct_most_phh_10_24, cid=id_drug, rid=id_genes)
  #means<- rowMeans(gct_drug@mat)
  medians <- apply(gct_drug@mat, 1, median) 
  heat_df[drug] <- medians
}
for (drug in no_concern_drugs){
  id_drug <- which(gct_no_phh_10_24@cdesc$pert_iname == drug)
  id_genes <- which(gct_no_phh_10_24@rid %in% gene_ids)
  gct_drug <- subset.gct(gct_no_phh_10_24, cid=id_drug, rid=id_genes)
  #means<- rowMeans(gct_drug@mat)
  medians <- apply(gct_drug@mat, 1, median) 
  heat_df[drug] <- medians
}
# Save the table
output_table <- "/home/quim/PHD/Projects/camda/results/gene_test_heatmap_data_phh_10_24_mixlessmost.tsv"
write.table(heat_df, file = output_table,row.names=TRUE, na="-",col.names=TRUE, sep="\t")

# Draw the heatmap
#heatmap(as.matrix(heat_df), Colv = NA, Rowv = NA, scale="column")
#heatmap.2(as.matrix(heat_df), col=bluered, Colv = NA, Rowv = NA, scale="row", trace="none")
#heatmap.2(as.matrix(heat_df), col=bluered, Colv = NA, Rowv = NA, scale="none", trace="none") #no scale means we use the raw z-scores
#cc <- c(rep("green", times=length(most_concern_drugs)), rep("yellow", times=length(no_concern_drugs)))
#cc2 <- c(rep("#004f00", times=length(most_concern_drugs)), rep("#9a9a00", times=length(no_concern_drugs)))
#heatmap.2(as.matrix(heat_df), col=bluered, Colv = NA, Rowv = NA, scale="none", trace="none", ColSideColors=cc, colCol=cc2) #no scale means we use the raw z-scores

type <- c(rep("Most-DILI-Concern", times=length(most_concern_drugs)), rep("No-DILI-Concern", times=length(no_concern_drugs)))
annotation <- HeatmapAnnotation(df = data.frame(type = type))
#Heatmap(heat_df, name = "expression", km = 0, col = colorRamp2(c(min(heat_df), 0, max(heat_df)), c("blue", "white", "red")), top_annotation = annotation, cluster_rows=FALSE, cluster_columns=FALSE)

# New heatmap using ComplexHeatmap
heat_df_for_plot <- heat_df
heat_df_for_plot <- apply(heat_df_for_plot, c(1,2),  function(x) {ifelse(abs(x)>1.5, x, 0)})

Cairo::CairoPDF("paths_all.pdf", width = 10, height = 10) # Save in PDF
Heatmap(heat_df_for_plot, name = "expression", km = 0, col = colorRamp2(c(min(heat_df), 0, max(heat_df)), c("blue", "white", "red")), top_annotation = annotation, cluster_rows=TRUE, cluster_columns=FALSE,  row_names_gp = gpar(fontsize = 7)  ,column_names_gp =  gpar(fontsize = 7)  )
dev.off()
```


## Perform the same analysis removing the outliers daunorubicin and vorinostat


```{r}
# Remove daunorubicin and vorinostat from no_concern_drugs
outliers = c('daunorubicin', 'vorinostat')
no_concern_drugs_noout <- no_concern_drugs[!no_concern_drugs %in% outliers]
id_no_phh_10_24_noout <- which(!gct_no_phh_10_24@cdesc$pert_iname %in% outliers)
gct_no_phh_10_24_noout <- subset.gct(gct_no_phh_10_24, cid=id_no_phh_10_24_noout)

# Create new dataframe to store the results
cols <- c("gene_id", "landmark", "statistic", "p.value")
wilcox_noout_df <- data.frame(matrix(ncol = length(cols), nrow=0))
colnames(wilcox_noout_df) <- cols

# Calculate the wilcoxon for each gene and store the results
#output_table <- "/home/quim/PHD/Projects/camda/results/gene_test_phh_10_24_noout.tsv"
output_table <- "/home/quim/PHD/Projects/camda/results/gene_test_landmark_phh_10_24_noout_mixlessmost.tsv"
if(!file.exists(output_table)){
  #for (gene_id in gct_most_phh_10_24@rid) {
  for (gene_id in landmark_genes) {
    most_expr <- gct_most_phh_10_24@mat[gct_most_phh_10_24@rid == gene_id]
    no_expr <- gct_no_phh_10_24_noout@mat[gct_no_phh_10_24_noout@rid == gene_id]
    res <- wilcox.test(most_expr, no_expr, alternative = "two.sided", paired = FALSE)
    landmark <- gene_id %in% landmark_genes
    wilcox_noout_df[nrow(wilcox_noout_df)+1,] <- c(gene_id, landmark, res$statistic, res$p.value)
  }
  wilcox_noout_df$p.value <- as.numeric(wilcox_noout_df$p.value)
  wilcox_noout_df$p.adjust <- p.adjust(wilcox_noout_df$p.value, "BH")
  write.table(wilcox_noout_df, file = output_table,row.names=FALSE, na="-",col.names=TRUE, sep="\t")
}else{
  wilcox_noout_df = read.csv(output_table, header = TRUE, sep = "\t")
}
head(wilcox_noout_df)
```


```{r}
# Define the genes found
all_wilcox_gene_ids_noout <- unique(wilcox_noout_df$gene_id[wilcox_noout_df$p.value<0.01])
all_wilcox_gene_ids_noout_05 <- unique(wilcox_noout_df$gene_id[wilcox_noout_df$p.value<0.05])
landmark_wilcox_gene_ids_noout <- unique(wilcox_noout_df$gene_id[wilcox_noout_df$p.value<0.01 & wilcox_noout_df$landmark==TRUE])
landmark_wilcox_gene_ids_noout_05 <- unique(wilcox_noout_df$gene_id[wilcox_noout_df$p.value<0.05 & wilcox_noout_df$landmark==TRUE])

# Get a vector of gene symbols ordered equal to the vector of gene IDs
genes_selected_noout <- landmark_wilcox_gene_ids_noout_05
genes_noout_df <- data.frame(1:length(genes_selected_noout), genes_selected_noout)
colnames(genes_noout_df)<- c("id", "pr_gene_id")
genes_noout_df <- merge(genes_noout_df, gene_info_df[,c("pr_gene_id", "pr_gene_symbol")], by="pr_gene_id")
genes_noout_df <- genes_noout_df[order(genes_noout_df$id), ]
genes_noout_df$id <- NULL

# Create new dataframe to store the matrix
cols <- c(most_concern_drugs, no_concern_drugs_noout)
rows <- genes_noout_df$pr_gene_symbol
heat_noout_df <- data.frame(matrix(ncol = length(cols), nrow=length(rows)))
colnames(heat_noout_df) <- cols
rownames(heat_noout_df) <- rows
gene_ids <- genes_noout_df$pr_gene_id

# Introduce the data in the matrix
for (drug in most_concern_drugs){
  id_drug <- which(gct_most_phh_10_24@cdesc$pert_iname == drug)
  id_genes <- which(gct_most_phh_10_24@rid %in% gene_ids)
  gct_drug <- subset.gct(gct_most_phh_10_24, cid=id_drug, rid=id_genes)
  #means<- rowMeans(gct_drug@mat)
  medians <- apply(gct_drug@mat, 1, median) 
  heat_noout_df[drug] <- medians
}
for (drug in no_concern_drugs_noout){
  id_drug <- which(gct_no_phh_10_24_noout@cdesc$pert_iname == drug)
  id_genes <- which(gct_no_phh_10_24_noout@rid %in% gene_ids)
  gct_drug <- subset.gct(gct_no_phh_10_24_noout, cid=id_drug, rid=id_genes)
  #means<- rowMeans(gct_drug@mat)
  medians <- apply(gct_drug@mat, 1, median) 
  heat_noout_df[drug] <- medians
}
# Save the table
output_table <- "/home/quim/PHD/Projects/camda/results/gene_test_heatmap_data_phh_10_24_noout.tsv"
write.table(heat_noout_df, file = output_table,row.names=TRUE, na="-",col.names=TRUE, sep="\t")

# ComplexHeatmap
type <- c(rep("Most-DILI-Concern", times=length(most_concern_drugs)), rep("No-DILI-Concern", times=length(no_concern_drugs_noout)))
annotation <- HeatmapAnnotation(df = data.frame(type = type))
Cairo::CairoPDF("/home/quim/PHD/Projects/camda/results/heatmap_phh_10_24_landmark_noout_mixlessmost.pdf", width = 10, height = 10) # Save in PDF
Heatmap(heat_noout_df, name = "expression", km = 0, col = colorRamp2(c(min(heat_noout_df), 0, max(heat_noout_df)), c("blue", "white", "red")), top_annotation = annotation, cluster_rows=TRUE, cluster_columns=FALSE,  row_names_gp = gpar(fontsize = 7)  ,column_names_gp =  gpar(fontsize = 7))
dev.off()

# ComplexHeatmap removing expression lower than abs(1.5)
heat_df_for_plot <- heat_noout_df
heat_df_for_plot <- apply(heat_df_for_plot, c(1,2),  function(x) {ifelse(abs(x)>1.5, x, 0)})
type <- c(rep("Most-DILI-Concern", times=length(most_concern_drugs)), rep("No-DILI-Concern", times=length(no_concern_drugs_noout)))
annotation <- HeatmapAnnotation(df = data.frame(type = type))
#Cairo::CairoPDF("/home/quim/PHD/Projects/camda/results/heatmap_phh_mixlessmost_10_24_landmark_noout_above1_5_mixlessmost.pdf", width = 10, height = 10) # Save in PDF
Cairo::CairoPNG("/home/quim/PHD/Projects/camda/results/heatmap_phh_mixlessmost_10_24_landmark_noout_above1_5_mixlessmost.png", width = 500, height = 500) # Save in PNG
Heatmap(heat_df_for_plot, name = "expression", km = 0, col = colorRamp2(c(min(heat_noout_df), 0, max(heat_noout_df)), c("blue", "white", "red")), top_annotation = annotation, cluster_rows=TRUE, cluster_columns=FALSE,  row_names_gp = gpar(fontsize = 7)  ,column_names_gp =  gpar(fontsize = 7)  )
dev.off()
```


